---
title: "Ch 2 Code (All)"
author: "Victoria Field"
date: "6/9/2020"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dataMaid)
library(dplyr)
library(tidyverse)
library(Matching)
library(MatchIt)
library(reshape2)
library(plyr)
library(ggplot2)
library(DHARMa)
library(magrittr)
library(pryr)
library(MASS)
library(car)
library(lme4)
require(lmerTest)
library(lattice)
library(predictmeans)
```

# Data Cleaning
```{r}
getwd()
CSLAP<-read.csv("./raw_data/CSLAP 2012-2017 EDI Original.csv", na.strings = c("", " ", "NA", "NaN"))
```

```{r}
str(CSLAP)
```

Based on output from `str()` and prior knowledge of the data, to clean this dataset I need to:

-Change column names and remove units 

-Identify duplicate levels of `Lake_Name`, we know there is a duplicate because we have 73 levels of the unique factor `LakeID` and 72 levels of `Lake_Name`

-Split `Sample_Date` variable into two more columns: `Sample_Year` and `Sample_Month`

-For each lake, identify how many years the observations span (If less than 3, remove from dataset)

-Clean levels of `Data_Provider` to one level ("CSL")

-Clean levels of `Info_Type`

-Remove character strings from `ESF_Microcystin_ug.L`

-Read in a merge values for Mean Depth, Catchment Area to Surface Area Ratio, Dreissenid Status By Year, and %Agricultural Cover


## Change column names 
```{r}
colnames(CSLAP)<-c("LakeID", "Lake_Name", "Proj_Code", "Sample_Name", "Sample_Date", "Data_Provider", "Info_Type", "LocationID", "Latitude", "Longitude", "Sample_Depth", "Secchi_Depth", "DO", "Bottom_DO","pH", "TP", "TDP", "Conductance", "Temp_Air", "Temp_Water", "TN", "TDN" , "Ammonia", "NOx", "Nitrate", "Nitrite", "True_Color", "Chloride", "Calcium", "Chl.a_FP", "Extracted_Chl.a", "UFI_SBK_FP_Chl.a", "UFI_SBK_FP_BGA", "UFI_SBK_Microcystin", "ESF_FP_Chl.a","ESF_FP_BGA", "ESF_Microcystin", "HAB_Status", "Alkalinity_CaCO3", "DOC", "UV_254", "Iron", "Manganese", "Arsenic", "QA", "QB", "QC")
```

## Change duplicate `Lake_Name`:"Forest Lake" and remove white space from `Lake_Name`: "Hadlock Pond"

We want the lake with `LakeID`:"1301FOR0441" to be called "Lake Forest" instead of "Forest Lake". And we need to change "Hadlock Pond " to "Hadlock Pond"
```{r}
CSLAP$Lake_Name<-as.character(CSLAP$Lake_Name) #first, make this column a character vector instead of factor
CSLAP[CSLAP$LakeID == "1301FOR0441", 2] <- rep("Lake Forest", 40) #then, replace the specified cells with "Lake Forest"
CSLAP[CSLAP$LakeID == "1005HAD0432", 2] <- rep("Hadlock Pond", 78) #then, replace the specified cells with "Hadlock Pond"
CSLAP$Lake_Name<-as.factor(CSLAP$Lake_Name) #last, change the column back to factor vector
```

## Create new columns for `Sample_Year` and `Sample_Month`
```{r}
#extract sample year and month 
CSLAP$Sample_Year<-as.numeric(format(as.Date(CSLAP$Sample_Date, format="%m/%d/%Y"),"%Y"))
CSLAP$Sample_Month<-as.numeric(format(as.Date(CSLAP$Sample_Date, format="%m/%d/%Y"),"%m"))
```

## For each lake, check how many years they were enrolled in CSLAP. Remove lakes with less than 3 years 
```{r}
table(CSLAP$LakeID, CSLAP$Sample_Year)

CSLAP$LakeID<-as.character(CSLAP$LakeID) #first, change to a character vector

CSLAP<-CSLAP[CSLAP$LakeID != "0403RUS0146", ]
CSLAP<-CSLAP[CSLAP$LakeID != "0704CAN0286", ] 
CSLAP<-CSLAP[CSLAP$LakeID != "0801EFF0426", ] 
CSLAP<-CSLAP[CSLAP$LakeID != "1104PAR0432", ] 
CSLAP<-CSLAP[CSLAP$LakeID != "0602PLY0107", ] #Plymouth reservoir will also be removed because it is eutrophic

CSLAP$LakeID<-as.factor(CSLAP$LakeID)
CSLAP$LakeID<-droplevels(CSLAP$LakeID)
levels(CSLAP$LakeID)
```

## Change levels of `Data_Provider`
```{r}
levels(CSLAP$Data_Provider)
CSLAP$Data_Provider<-as.character(CSLAP$Data_Provider)
CSLAP[CSLAP$Data_Provider == "csl", 6] <- "CSL"
CSLAP$Data_Provider<-as.factor(CSLAP$Data_Provider)
CSLAP$Data_Provider<-droplevels(CSLAP$Data_Provider)
levels(CSLAP$Data_Provider)
```

## Change levels of `Info_Type`
```{r}
levels(CSLAP$Info_Type)
CSLAP$Info_Type<-as.character(CSLAP$Info_Type)
CSLAP[CSLAP$Info_Type == "ow", 7] <- "OW"
CSLAP[CSLAP$Info_Type == "sb", 7] <- "SB"
CSLAP$Info_Type<-as.factor(CSLAP$Info_Type)
CSLAP$Info_Type<-droplevels(CSLAP$Info_Type)
levels(CSLAP$Info_Type)
```

## Clean `ESF_Microcystin` vector
```{r}
# ESF_Microcystin should be a numeric column,but it's listed as factor. We need to remove character values from this column 

CSLAP$ESF_Microcystin[CSLAP$ESF_Microcystin == "ND"] <- NA
CSLAP$ESF_Microcystin[CSLAP$ESF_Microcystin == "no filter"] <- NA

CSLAP$ESF_Microcystin <- as.numeric(as.character(CSLAP$ESF_Microcystin))

str(CSLAP$ESF_Microcystin)
```

## Add TN:TP column 
```{r}
CSLAP$TN_TP<-CSLAP$TN/CSLAP$TP
```

## Check for missing values 
```{r}
result <- lapply(CSLAP, identifyMissing)
result
```
```{r}
CSLAP$TN_TP[CSLAP$TN_TP == "Inf"] <- NA
CSLAP$TN_TP[CSLAP$TN_TP == "NaN"] <- NA

result <- lapply(CSLAP, identifyMissing)
result
```

## Read in and merge mean depth, CA:SA, dreissenids by year, and %ag
```{r}
Dreissenids<-read.csv("./raw_data/Dreissenid_Status_byYear.csv", header=TRUE)
Dreissenids$Lake_Name<-as.character(Dreissenids$Lake_Name) 
Dreissenids[200:204,2] <- rep("Hadlock Pond",5)
Dreissenids$Lake_Name<-as.factor(Dreissenids$Lake_Name)

CASA_MD<-read.csv("./raw_data/CASA_Mean_Depth.csv", header=TRUE)
CASA_MD$Lake_Name<-as.character(CASA_MD$Lake_Name) 
CASA_MD[27,2] <- "Hadlock Pond"
CASA_MD$Lake_Name<-as.factor(CASA_MD$Lake_Name)

Ag<-read.csv("./raw_data/Percent Ag Cover.csv", header=TRUE)
Ag$Lake_Name<-as.character(Ag$Lake_Name) 
Ag[27,1] <- "Hadlock Pond"
Ag$Lake_Name<-as.factor(Ag$Lake_Name)

CSLAP<-merge(CSLAP, Dreissenids, by=c("LakeID", "Lake_Name", "Sample_Year"), all.x =TRUE)
CSLAP<-merge(CSLAP, CASA_MD, by=c("LakeID", "Lake_Name"), all.x=TRUE)
CSLAP<-merge(CSLAP, Ag, by="Lake_Name", all.x=TRUE)
```

Now we have a clean, full dataset. Next, we need to create the reduced dataset using propensity score matching

## Create Reduced Dataset 

### Read in lake data
```{r}
Lake_List<-read.csv("./raw_data/LakeList.csv", header=TRUE)
Lake_List<-merge(Lake_List, Ag, by="Lake_Name", na.rm=FALSE, all.x=TRUE) #merge %ag
Lake_List<-Lake_List[Lake_List$LakeID != "0403RUS0146",] #Remove Lake Rushford
```

### Propensity Score Estimation 

```{r}
m_ps <- glm(Dreissenids ~ Mean_Depth_m + CA.SA + Percent_Ag,
            family = binomial(), data = Lake_List)
summary(m_ps)
```

```{r}
prs_df <- data.frame(pr_score = predict(m_ps, type = "response"),
                     Dreissenids = m_ps$model$Dreissenids)
head(prs_df)
```

```{r}
labs <- paste("Invasion Status:", c("Invaded", "Uninvaded"))
prs_df %>%
  mutate(Dreissenids = ifelse(Dreissenids == 1, labs[1], labs[2])) %>%
  ggplot(aes(x = pr_score)) +
  geom_histogram(color = "white") +
  facet_wrap(~Dreissenids) +
  xlab("Probability of being invaded") +
  theme_bw()
```

```{r}
lakes_nomiss <- Lake_List %>%  # MatchIt does not allow missing values
  dplyr::select(CA.SA, Mean_Depth_m, Percent_Ag, LakeID, Lake_Name, Dreissenids) %>%
  na.omit()

mod_match <- matchit(Dreissenids ~ CA.SA + Mean_Depth_m + Percent_Ag,
                     method = "nearest", data = lakes_nomiss)

df_match <- match.data(mod_match)
```

## Extract observations that come from lakes in the reduced set to create new, clean reduced dataset

```{r}
redLakes<-df_match[,c(4,6)]

redCSLAP<-merge(CSLAP, redLakes, by="LakeID")

redCSLAP<-redCSLAP[,1:54] #removed redundant Dreissenids column

colnames(redCSLAP)[51] <- "Dreissenids"
```

# Data Summaries and Histograms 

## Global Dataset 

### Data Summaries
```{r}
#melt data frame by Dreissenids and Info_type so we can summarize all our variables of interest in one code chunk 
CSLAP2<-CSLAP[, c(8, 13, 17, 21, 22, 28, 32, 36, 37, 38, 50:54)]
meltCSLAP <- melt(CSLAP2, id=c("Dreissenids", "Info_Type"), na.rm=TRUE)

#summarize each variable split by invasion status and info
VariableSummaries<-ddply(meltCSLAP, c("Dreissenids", "Info_Type", "variable"), summarize,
      Mean = mean(value, na.rm=TRUE), 
      Median = median(value, na.rm=TRUE), 
      Min = min(value, na.rm=TRUE), 
      Max = max(value, na.rm=TRUE), 
      SD = sd(value, na.rm=TRUE))
VariableSummaries

#I will pull out only the values that I need to make copying to a word document "easy"
VariableSummaries[c(1, 6:11,15:18, 20:22, 27, 38:34, 47:50, 55:57),]

#Write this dataframe to a csv so the values can be easily put into a word document 
write.csv(VariableSummaries, "./output_data/VariableSummaries-June2020.csv")
```

### Histograms

First we will split our dataset based on information type
```{r}
OWCSLAP<-CSLAP[CSLAP$Info_Type == "OW",]
BSCSLAP<-CSLAP[CSLAP$Info_Type == "BS",]
SBCSLAP<-CSLAP[CSLAP$Info_Type == "SB",]
```

#### Open Water TP
```{r}
ggplot(OWCSLAP, aes(x=TP)) +
  geom_histogram(position="identity", color="black", fill="white")+
  facet_wrap(~Dreissenids)+ 
  geom_density()+
  theme_minimal()+
  labs(title="Open water TP", x="TP (mg/L)", y="Frequency")
```

#### Open Water TN
```{r}
ggplot(OWCSLAP, aes(x=TN)) +
  geom_histogram(position="identity", color="black", fill="white")+
  facet_wrap(~Dreissenids)+ 
  geom_density()+
  theme_minimal()+
  labs(title="Open water TN", x="TN (mg/L)", y="Frequency")
```

#### Open Water TN:TP
```{r}
ggplot(OWCSLAP, aes(x=TN_TP)) +
  geom_histogram(position="identity", color="black", fill="white")+
  facet_wrap(~Dreissenids)+ 
  geom_density()+
  theme_minimal()+
  labs(title="Open water TN:TP", x="TN:TP", y="Frequency")+
  scale_x_continuous(limits=c(0,100))
```

#### Open Water Chlorophyll-a
```{r}
ggplot(OWCSLAP, aes(x=Extracted_Chl.a)) +
  geom_histogram(position="identity", color="black", fill="white")+
  facet_wrap(~Dreissenids)+ 
  geom_density()+
  theme_minimal()+
  labs(title="Open water chlorophyll a", x="Chlorophyll a (ug/L)", y="Frequency")
```

#### Secchi Depth
```{r}
ggplot(OWCSLAP, aes(x=Secchi_Depth)) +
  geom_histogram(position="identity", color="black", fill="white")+
  facet_wrap(~Dreissenids)+ 
  geom_density()+
  theme_minimal()+
  labs(title="Secchi Depth", x="Secchi depth (m)", y="Frequency")
```

#### True Color
```{r}
ggplot(OWCSLAP, aes(x=True_Color)) +
  geom_histogram(position="identity", color="black", fill="white")+
  facet_wrap(~Dreissenids)+ 
  geom_density()+
  theme_minimal()+
  labs(title="True Color", x="True color (PTU)", y="Frequency")
```

#### Shoreline Bloom Chlorophyll
```{r}
ggplot(SBCSLAP, aes(x=ESF_FP_Chl.a)) +
  geom_histogram(position="identity", color="black", fill="white")+
  facet_wrap(~Dreissenids)+ 
  geom_density()+
  theme_minimal()+
  labs(title="Shoreline Bloom Chlorophyll a", x="Chlorophyll a (ug/L)", y="Frequency")
```

#### Shoreline Bloom Microcystin
```{r}
ggplot(SBCSLAP, aes(x=ESF_Microcystin)) +
  geom_histogram(position="identity", color="black", fill="white")+
  facet_wrap(~Dreissenids)+ 
  geom_density()+
  theme_minimal()+
  labs(title="Shoreline Bloom Microcystin", x="Microcystin (ug/L)", y="Frequency")
```

#### Mean Depth 
```{r}
#Quickly change factor level labels for Dreissenids
Lake_List[Lake_List$Dreissenids == 0, 5] <- rep("Uninvaded", 60)
Lake_List[Lake_List$Dreissenids == 1, 5] <- rep("Invaded", 8)
```

```{r}
ggplot(Lake_List, aes(x=Mean_Depth_m)) +
  geom_histogram(position="identity", color="black", fill="white")+
  facet_wrap(~Dreissenids)+ 
  geom_density()+
  theme_minimal()+
  labs(title="Mean Depth", x="Mean depth (m)", y="Frequency")
```

#### CA:SA

```{r}
ggplot(Lake_List, aes(x=CA.SA)) +
  geom_histogram(position="identity", color="black", fill="white")+
  facet_wrap(~Dreissenids)+ 
  geom_density()+
  theme_minimal()+
  labs(title="Catchment Area to Surface Area Ratio", x="CA:SA", y="Frequency")
```

#### Percent Agriculture

```{r}
ggplot(Lake_List, aes(x=Percent_Ag)) +
  geom_histogram(position="identity", color="black", fill="white")+
  facet_wrap(~Dreissenids)+ 
  geom_density()+
  theme_minimal()+
  labs(title="Percent Agricultural Cover in Watershed", x="Agriculture (%)", y="Frequency")
```

## Reduced Dataset 

### Data Summaries
```{r}
#melt data frame by Dreissenids and Info_type so we can summarize all our variables of interest in one code chunk 
redCSLAP2<-redCSLAP[, c(8, 13, 17, 21, 22, 28, 32, 36, 37, 38, 50:54)]
redmeltCSLAP <- melt(redCSLAP2, id=c("Dreissenids", "Info_Type"), na.rm=TRUE)

#summarize each variable split by invasion status and info
redVariableSummaries<-ddply(meltCSLAP, c("Dreissenids", "Info_Type", "variable"), summarize,
      Mean = mean(value, na.rm=TRUE), 
      Median = median(value, na.rm=TRUE), 
      Min = min(value, na.rm=TRUE), 
      Max = max(value, na.rm=TRUE), 
      SD = sd(value, na.rm=TRUE))
redVariableSummaries

#Write this dataframe to a csv so the values can be easily put into a word document 
write.csv(redVariableSummaries, "./output_data/redVariableSummaries-June2020.csv")
```

### Histograms

First we will split our dataset based on information type
```{r}
redOWCSLAP<-redCSLAP[redCSLAP$Info_Type == "OW",]
redBSCSLAP<-redCSLAP[redCSLAP$Info_Type == "BS",]
redSBCSLAP<-redCSLAP[redCSLAP$Info_Type == "SB",]
```

#### Open Water TP
```{r}
ggplot(redOWCSLAP, aes(x=TP)) +
  geom_histogram(position="identity", color="black", fill="white")+
  facet_wrap(~Dreissenids)+ 
  geom_density()+
  theme_minimal()+
  labs(title="Open water TP", x="TP (mg/L)", y="Frequency")
```

#### Open Water TN
```{r}
ggplot(redOWCSLAP, aes(x=TN)) +
  geom_histogram(position="identity", color="black", fill="white")+
  facet_wrap(~Dreissenids)+ 
  geom_density()+
  theme_minimal()+
  labs(title="Open water TN", x="TN (mg/L)", y="Frequency")
```

#### Open Water TN:TP
```{r}
ggplot(redOWCSLAP, aes(x=TN_TP)) +
  geom_histogram(position="identity", color="black", fill="white")+
  facet_wrap(~Dreissenids)+ 
  geom_density()+
  theme_minimal()+
  labs(title="Open water TN:TP", x="TN:TP", y="Frequency")+
  scale_x_continuous(limits=c(0,100))
```

#### Open Water Chlorophyll-a
```{r}
ggplot(redOWCSLAP, aes(x=Extracted_Chl.a)) +
  geom_histogram(position="identity", color="black", fill="white")+
  facet_wrap(~Dreissenids)+ 
  geom_density()+
  theme_minimal()+
  labs(title="Open water chlorophyll a", x="Chlorophyll a (ug/L)", y="Frequency")
```

#### Secchi Depth
```{r}
ggplot(redOWCSLAP, aes(x=Secchi_Depth)) +
  geom_histogram(position="identity", color="black", fill="white")+
  facet_wrap(~Dreissenids)+ 
  geom_density()+
  theme_minimal()+
  labs(title="Secchi Depth", x="Secchi depth (m)", y="Frequency")
```

#### True Color
```{r}
ggplot(redOWCSLAP, aes(x=True_Color)) +
  geom_histogram(position="identity", color="black", fill="white")+
  facet_wrap(~Dreissenids)+ 
  geom_density()+
  theme_minimal()+
  labs(title="True Color", x="True color (PTU)", y="Frequency")
```

#### Shoreline Bloom Chlorophyll
```{r}
ggplot(redSBCSLAP, aes(x=ESF_FP_Chl.a)) +
  geom_histogram(position="identity", color="black", fill="white")+
  facet_wrap(~Dreissenids)+ 
  geom_density()+
  theme_minimal()+
  labs(title="Shoreline Bloom Chlorophyll a", x="Chlorophyll a (ug/L)", y="Frequency")
```

#### Shoreline Bloom Microcystin
```{r}
ggplot(redSBCSLAP, aes(x=ESF_Microcystin)) +
  geom_histogram(position="identity", color="black", fill="white")+
  facet_wrap(~Dreissenids)+ 
  geom_density()+
  theme_minimal()+
  labs(title="Shoreline Bloom Microcystin", x="Microcystin (ug/L)", y="Frequency")
```

# Historical Analyses for Invaded Lakes

## Secchi Depth
Because of the way the data were given to me, Secchi depth has to be done first on its own
```{r}
hisCSLAP<-read.csv("./historical_data/Historical_CSLAP.csv", na.strings=c("", " ", "ND", "NA", "no filter", "nofilter"))

#extract sample year and month 
hisCSLAP$Sample_Year<-as.numeric(format(as.Date(hisCSLAP$Sample_Date, format="%m/%d/%Y"),"%Y"))
hisCSLAP$Sample_Month<-as.numeric(format(as.Date(hisCSLAP$Sample_Date, format="%m/%d/%Y"),"%m"))

#read in invasion year information 
Invasion_Year<-read.csv("./raw_data/InvasionYear.csv", header=TRUE, colClasses=c("factor", "numeric"))

#merge with hisCSLAP
hisCSLAP<-merge(hisCSLAP, Invasion_Year, all.x=TRUE)

#create new column `Years_Invaded` by subtracting `Invasion Year` from `Sample Year`
hisCSLAP$Years_Invaded<-hisCSLAP$Sample_Year-hisCSLAP$Invasion_Year

#turn `Sample_Year`, `Sample_Month`, and `Invasion_Year` back to factors
hisCSLAP$Sample_Year<-as.factor(hisCSLAP$Sample_Year)
hisCSLAP$Sample_Month<-as.factor(hisCSLAP$Sample_Month)
hisCSLAP$Invasion_Year<-as.factor(hisCSLAP$Invasion_Year)

#create column with 'Invaded' or "Uninvaded" based on `Years Invaded`
hisCSLAP$Dreissenids<-ifelse(hisCSLAP$Years_Invaded >= 0, "Invaded", "Uninvaded")
hisCSLAP$Dreissenids<-as.factor(hisCSLAP$Dreissenids)
```

```{r}
#Aggregating yearly means by lake and years invaded
Secchi<-aggregate(Secchi_Depth_m ~ Lake_Name + Years_Invaded, data=hisCSLAP, FUN=mean)

Chl<-aggregate(Extracted_Chl_ug.L ~ Lake_Name + Years_Invaded, data=hisCSLAP, FUN=mean)
```

```{r}
#t-test pre and post invasion and MEMs
hist(hisCSLAP$Secchi_Depth_m)
t.test(Secchi_Depth_m~Dreissenids, data=hisCSLAP)

Secchi.lmer<-lmer(Secchi_Depth_m ~ Dreissenids + (1|LakeID), data=hisCSLAP)

summary(Secchi.lmer)
```

```{r}
ggplot(hisCSLAP, aes(x=Dreissenids, y=Secchi_Depth_m))+
  geom_boxplot()+
  theme_minimal()+
  labs(y = "Secchi Depth (m)")

#ggsave("historical_Secchi.png")
```

## Data Read In For the Rest of the Variables
```{r}
#historicalCSLAP df edited to match DeRuyter and EatonBrook dfs
hisCSLAP<-read.csv("./historical_data/Historical_CSLAP_edited.csv", na.strings=c("", " ", "ND", "NA", "no filter", "nofilter"))

#DeRuyter and EatonBrook dfs
Eaton<-read.csv("./historical_data/EatonHistoricalClean2.csv", na.strings=c("", " ", "ND", "NA", "no filter", "nofilter"))
Der<-read.csv("./historical_data/DeruyterHistoricalClean2.csv", na.strings=c("", " ", "ND", "NA", "no filter", "nofilter"))

#change levels of Info_Type 
hisCSLAP$Info_Type[hisCSLAP$Info_Type == "ow"]<-"OW"
hisCSLAP$Info_Type[hisCSLAP$Info_Type == "sb"]<-"SB"
hisCSLAP$Info_Type<-as.factor(hisCSLAP$Info_Type)
hisCSLAP$Info_Type<-droplevels(hisCSLAP$Info_Type)
```

```{r}
#Merge hisCSLAP with Deruyter and Eatonbrook 

#remove any observations from Eaton or DeRuyter from hisCSLAP 
hisCSLAP<-hisCSLAP[hisCSLAP$Lake_Name != "De Ruyter Reservoir",]
hisCSLAP<-hisCSLAP[hisCSLAP$Lake_Name != "Eaton Brook Reservoir",]

#row bind cleaned DeRuyter and EatonBrook 
hisCSLAP<-rbind(hisCSLAP, Eaton)
hisCSLAP<-rbind(hisCSLAP, Der)
```

## Further organize hisCSLAP, and merge physical characteristics and invasion status 
```{r}
#extract sample year and month 
hisCSLAP$Sample_Year<-as.numeric(format(as.Date(hisCSLAP$Sample_Date, format="%m/%d/%Y"),"%Y"))
hisCSLAP$Sample_Month<-as.numeric(format(as.Date(hisCSLAP$Sample_Date, format="%m/%d/%Y"),"%m"))

#merge Invasion Year with hisCSLAP
hisCSLAP<-merge(hisCSLAP, Invasion_Year, all.x=TRUE)

#create new column `Years_Invaded` by subtracting `Invasion Year` from `Sample Year`
hisCSLAP$Years_Invaded<-hisCSLAP$Sample_Year-hisCSLAP$Invasion_Year

#turn `Sample_Year`, `Sample_Month`, and `Invasion_Year` back to factors
hisCSLAP$Sample_Year<-as.factor(hisCSLAP$Sample_Year)
hisCSLAP$Sample_Month<-as.factor(hisCSLAP$Sample_Month)
hisCSLAP$Invasion_Year<-as.factor(hisCSLAP$Invasion_Year)

#create column with 'Invaded' or "Uninvaded" based on `Years Invaded`
hisCSLAP$Dreissenids<-ifelse(hisCSLAP$Years_Invaded >= 0, "Invaded", "Uninvaded")
hisCSLAP$Dreissenids<-as.factor(hisCSLAP$Dreissenids)

#we will also remove and shoreline bloom samples 
hisCSLAP<-hisCSLAP[hisCSLAP$Info_Type != "SB",]
hisCSLAP$Info_Type<-droplevels(hisCSLAP$Info_Type)

#create a TN:TP column 
hisCSLAP$TN_TP<-hisCSLAP$TN_mg.L/hisCSLAP$TP_mg.L

#finally, create a df with only OW samples and a df with only BS samples 
OWhisCSLAP<-hisCSLAP[hisCSLAP$Info_Type == "OW",]
BShisCSLAP<-hisCSLAP[hisCSLAP$Info_Type == "BS",]
```

## Aggregating yearly means by lake and years invaded
```{r}
Chl<-aggregate(Extracted_Chl_ug.L ~ Lake_Name + Years_Invaded, data=hisCSLAP[hisCSLAP$Info_Type == "OW",], FUN=mean)

OWTP<-aggregate(TP_mg.L ~ Lake_Name + Lake_Name + Years_Invaded, data=hisCSLAP[hisCSLAP$Info_Type == "OW",], FUN=mean)

BSTP<-aggregate(TP_mg.L ~ Lake_Name + Lake_Name + Years_Invaded, data=hisCSLAP[hisCSLAP$Info_Type == "BS",], FUN=mean)

OWTN<-aggregate(TN_mg.L ~ Lake_Name + Lake_Name + Years_Invaded, data=hisCSLAP[hisCSLAP$Info_Type == "OW",], FUN=mean)

OWTN_TP<-aggregate(TN_TP ~ Lake_Name + Lake_Name + Years_Invaded, data=hisCSLAP[hisCSLAP$Info_Type == "OW",], FUN=mean)

TC<-aggregate(True_Color_PTU ~ Lake_Name + Years_Invaded, data=hisCSLAP[hisCSLAP$Info_Type == "OW",], FUN=mean)

Calc<-aggregate(Calcium_mg.L ~ Lake_Name + Years_Invaded, data=hisCSLAP[hisCSLAP$Info_Type == "OW",], FUN=mean)
```

## How do lakes change after invasion? (t-test for pre and post invasion, then MEMs)

### Extracted Chlorophyll-a
```{r}
hist(log(OWhisCSLAP$Extracted_Chl_ug.L))

t.test(Extracted_Chl_ug.L ~ Dreissenids, data=hisCSLAP[hisCSLAP$Info_Type == "OW",])

Chl.lmer<-lmer(log(Extracted_Chl_ug.L) ~ Dreissenids + (1|LakeID), data=hisCSLAP[hisCSLAP$Info_Type == "OW",])

summary(Chl.lmer)

plot(Chl.lmer)
```

### Open Water TP
```{r}
hist(log(OWhisCSLAP$TP_mg.L))

t.test(TP_mg.L ~ Dreissenids, data=hisCSLAP[hisCSLAP$Info_Type == "OW",])

OWTP.lmer<-lmer(log(TP_mg.L) ~ Dreissenids  + (1|LakeID), data=hisCSLAP[hisCSLAP$Info_Type == "OW",])

summary(OWTP.lmer)

plot(OWTP.lmer)
```

### Bottom Sample TP
```{r}
hist(log(BShisCSLAP$TP_mg.L))

t.test(TP_mg.L ~ Dreissenids, data=hisCSLAP[hisCSLAP$Info_Type == "BS",])

BSTP.lmer<-lmer(log(TP_mg.L) ~ Dreissenids + (1|LakeID), data=hisCSLAP[hisCSLAP$Info_Type == "BS",])

summary(BSTP.lmer)

plot(BSTP.lmer)
```

### Open Water TN
```{r}
hist(log(OWhisCSLAP$TN_mg.L))

t.test(TN_mg.L ~ Dreissenids, data=hisCSLAP[hisCSLAP$Info_Type == "OW",])

OWTN.lmer<-lmer(log(TN_mg.L) ~ Dreissenids + (1|LakeID), data=hisCSLAP[hisCSLAP$Info_Type == "OW",])

summary(OWTN.lmer)
```

### Open Water TN:TP
```{r}
hist(log(OWhisCSLAP$TN_TP))

t.test(TN_TP ~ Dreissenids, data=hisCSLAP[hisCSLAP$Info_Type == "OW",])

OWTN_TP.lmer<-lmer(log(TN_TP) ~ Dreissenids + (1|LakeID), data=hisCSLAP[hisCSLAP$Info_Type == "OW",])

summary(OWTN_TP.lmer)
```

### Open Water True Color
```{r}
t.test(True_Color_PTU ~ Dreissenids, data=hisCSLAP[hisCSLAP$Info_Type == "OW",])

TC.lmer<-lmer(True_Color_PTU ~ Dreissenids + (1|LakeID), data=hisCSLAP[hisCSLAP$Info_Type == "OW",])

summary(TC.lmer)
```

### Calcium
```{r}
t.test(Calcium_mg.L ~ Dreissenids, data=hisCSLAP[hisCSLAP$Info_Type == "OW",])

Calc.lmer<-lmer(Calcium_mg.L ~ Dreissenids + (1|LakeID), data=hisCSLAP[hisCSLAP$Info_Type == "OW",])

summary(Calc.lmer)
```

## Boxplots

### Chlorophyll-a
```{r}
ggplot(hisCSLAP[hisCSLAP$Info_Type == "OW",], aes(x=Dreissenids, y=Extracted_Chl_ug.L)) + 
  geom_boxplot()+
  theme_bw()+
  labs(y = "Extracted Chlorophyll-a (mg/L)")

#ggsave("./output_figures/historical_Chl.png")
```

### OWTP
```{r}
ggplot(hisCSLAP[hisCSLAP$Info_Type == "OW",], aes(x=Dreissenids, y=TP_mg.L)) + 
  geom_boxplot()+
  theme_bw()+
  labs(y = "Open Water TP (mg/L)")

#ggsave("./output_figures/historical_OWTP.png")
```

### BSTP
```{r}
ggplot(hisCSLAP[hisCSLAP$Info_Type == "BS",], aes(x=Dreissenids, y=TP_mg.L)) + 
  geom_boxplot()+
  theme_bw()+
  labs(y = "Bottom Sample TP (mg/L)")

#ggsave("./output_figures/historical_BSTP.png")
```

### OWTN
```{r}
ggplot(hisCSLAP[hisCSLAP$Info_Type == "OW",], aes(x=Dreissenids, y=TN_mg.L)) + 
  geom_boxplot()+
  theme_bw()+
  labs(y = "Open Water TN (mg/L)")

#ggsave("./output_figures/historical_OWTN.png")
```

### OWTN_TP
```{r}
ggplot(hisCSLAP[hisCSLAP$Info_Type == "OW",], aes(x=Dreissenids, y=TN_TP)) + 
  geom_boxplot()+
  theme_bw()+
  labs(y = "Open Water TN:TP")

#ggsave("./output_figures/historical_OWTN_TP.png")
```

### OW TC
```{r}
ggplot(hisCSLAP[hisCSLAP$Info_Type == "OW",], aes(x=Dreissenids, y=True_Color_PTU)) + 
  geom_boxplot()+
  theme_bw()+
  labs(y = "Open Water True Color")

#ggsave("./output_figures/historical_OWTC.png")
```

### OW Calcium
```{r}
ggplot(hisCSLAP[hisCSLAP$Info_Type == "OW",], aes(x=Dreissenids, y=Calcium_mg.L)) + 
  geom_boxplot()+
  theme_bw()+
  labs(y = "Open Water Calcium")

#ggsave("./output_figures/historical_OWCalcium.png")
```

# Mixed Effects Models

## Global Dataset 

### Open water TN:TP
```{r}
TNTP<-lmer(log(TN_TP+.01) ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m) + Percent_Ag +  (1|Sample_Year) + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=OWCSLAP)

summary(TNTP)

plot(TNTP)

qqPlot(resid(TNTP))
```

### Bottom Sample TP
```{r}
BSTP<-lmer(log(TP+.01) ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m) + Percent_Ag +  (1|Sample_Year) + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=BSCSLAP)

summary(BSTP)

plot(BSTP)

qqPlot(resid(BSTP))
```

### Open water TN

Full model error: `Model failed to converge`

-Remove `Sample_Month`. `Sample_Year`, `Sample_Year:LakeID` (low variance) to resolve error
```{r}
TN<-lmer(log(TN+.01) ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m) + Percent_Ag +   (1|LakeID), data=OWCSLAP)

summary(TN)

plot(TN)

qqPlot(resid(TN))
```

### Open water chlorophyll
```{r}
Chl<-lmer(log(Extracted_Chl.a+.01) ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m) + Percent_Ag +  (1|Sample_Year) + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=OWCSLAP)

summary(Chl)

plot(Chl)

qqPlot(resid(Chl))
```

### Secchi
```{r}
Secchi<-lmer(log(Secchi_Depth+.01) ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m) + Percent_Ag +  (1|Sample_Year) + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=OWCSLAP)

summary(Secchi)

plot(Secchi)

qqPlot(resid(Secchi))
```

### Open Water True Color
```{r}
TC<-lmer(log(True_Color+.01) ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m) + Percent_Ag +  (1|Sample_Year) + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=OWCSLAP)

summary(TC)

plot(TC)

qqPlot(resid(TC))
```

### Shoreline Bloom Chlorophyll a

Full model error: `boundary (singular) fit` and `model failed to converge`

Remove `Sample_Year`, `Sample_Month`, and `Sample_Year:LakeID` (low variance) to resolve
```{r}
SBChl<-lmer(log(ESF_FP_Chl.a+.01) ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m) + Percent_Ag  + (1|LakeID), data=SBCSLAP)

summary(SBChl)

plot(SBChl)

qqPlot(resid(SBChl))
```

### Shoreline Bloom Microcystin

We want to add TN:TP as a predictor (fixed effect), but these variables aren't collected with shoreline bloom data. So we will instead use the yearly average for a lake.
```{r, warning=FALSE, message=FALSE}
#Extracting average annual TN_TP for each lake 
library(plyr)
avgTNTP<-ddply(OWCSLAP, c("LakeID", "Sample_Year"), summarize, 
             Mean = mean(TN_TP, na.rm=TRUE))
colnames(avgTNTP)[colnames(avgTNTP)=="Mean"] <- "TN_TP"

#Merge these values to the SBCSLAP df 
SBCSLAP<-SBCSLAP[,c(1:49, 51:54)]
SBCSLAP<-merge(SBCSLAP, avgTNTP, by=c("LakeID", "Sample_Year"), all.x=TRUE, all.y=FALSE)
```

Full model error: `boundary (singular) fit` 

Remove `Sample_Month` and `Sample_Year:LakeID` to resolve
```{r}
SBmicro<-lmer(log(ESF_Microcystin) ~ Dreissenids + TN_TP + log(CA.SA) + Mean_Depth_m + Percent_Ag  + (1|Sample_Year) + (1|LakeID), data=SBCSLAP)

summary(SBmicro)

plot(SBmicro)

qqPlot(resid(SBmicro))
```

## Reduced Dataset

### Open water TN:TP
```{r}
redTNTP<-lmer(log(TN_TP+.01) ~ Dreissenids +  (1|Sample_Year) + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=redOWCSLAP)

summary(redTNTP)

plot(redTNTP)

qqPlot(resid(redTNTP))
```

### Bottom Sample TP
```{r}
redBSTP<-lmer(log(TP+.01) ~ Dreissenids  +  (1|Sample_Year) + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=redBSCSLAP)

summary(redBSTP)

plot(redBSTP)

qqPlot(resid(redBSTP))
```

### Open Water Chlorophyll-a

```{r}
redChl<-lmer(log(Extracted_Chl.a+1) ~ Dreissenids + (1|LakeID) + (1|Sample_Year) + (1|Sample_Month) + (1|Sample_Year:LakeID), data=redOWCSLAP)

summary(redChl)

plot(redChl)

qqPlot(resid(redChl))
```

### Open Water Secchi Depth

-`Sample_Year` and `Sample_Month` removed for having low variance
```{r}
redSecchi <- lmer(log(Secchi_Depth) ~ Dreissenids  + (1|LakeID)+ (1|Sample_Year) + (1|Sample_Month) + (1|Sample_Year:LakeID), data=redOWCSLAP)

summary(redSecchi)

plot(redSecchi)

qqPlot(resid(redSecchi))
```

### Open Water True Color 

Full model error: `Model failed to converge`

Remove `Sample_Year` and `Sample_Month` (low variance) to resolve
```{r}
redOWTC<-lmer(True_Color ~ Dreissenids  + (1|LakeID) + + (1|Sample_Year:LakeID), data=redOWCSLAP)

summary(redOWTC)

plot(redOWTC)

qqPlot(resid(redOWTC))
```

### Shoreline Bloom Chlorophyll a 

-`Sample_Year` removed for low variance

```{r}
redSBChl<-lmer(log(ESF_FP_Chl.a) ~ Dreissenids + (1|Sample_Month) + (1|LakeID) , data=redSBCSLAP)

summary(redSBChl)

plot(redSBChl)

qqPlot(resid(redSBChl))
```

## Shoreline Bloom Microcystin

Same as for the global dataset, we need to use average yearly TN:TP values
```{r, warning=FALSE}
#Extracting average annual TP for each lake 
library(plyr)
redavgTNTP<-ddply(redOWCSLAP, c("LakeID", "Sample_Year"), summarize, 
             Mean = mean(TN_TP, na.rm=TRUE))
colnames(redavgTNTP)[colnames(redavgTNTP)=="Mean"] <- "TN_TP"

#Merge these values to the SBCSLAP df 
redSBCSLAP<-redSBCSLAP[,c(1:49, 51:54)]
redSBCSLAP<-merge(redSBCSLAP, redavgTNTP, by=c("LakeID", "Sample_Year"), all.x=TRUE, all.y=FALSE)
```

Full model error: `failed to converge` 

Remove `Sample_Month`, `LakeID`,  and `Sample_Year:LakeID` to resolve
```{r}
redSBmicro<-lmer(log(ESF_Microcystin) ~ Dreissenids + log(TN_TP)  + (1|Sample_Year)  , data=redSBCSLAP)

summary(redSBmicro)

plot(redSBmicro)

qqPlot(resid(redSBmicro))
```

# Linear Regressions

## Global Dataset
```{r}
#Extracting average annual TP for each lake 
library(plyr)
avgTP<-ddply(OWCSLAP, c("LakeID", "Sample_Year"), summarize, 
             Mean = mean(TP, na.rm=TRUE))

#Taking only rows where there is a value for Microcystin 
Bloom<-CSLAP[!is.na(CSLAP$ESF_Microcystin),]

#Which rows are missing TP values? 
noTPbloom<-Bloom[is.na(Bloom$TP),]

#merge average TP to noTPbloom data frame 
avgTPbloom<-merge(noTPbloom, avgTP, by=c("LakeID", "Sample_Year"), all.x=TRUE, all.y=FALSE)

#remove original TP column, and re-name the average TP column 
avgTPbloom<-avgTPbloom[, c(1:16, 18:55)]
colnames(avgTPbloom)[colnames(avgTPbloom)=="Mean"] <- "TP"

#re-order columns 
avgTPbloom<-avgTPbloom[, c(1:16,54,17:53)]

#combine two data frames
hadTP<-Bloom[!is.na(Bloom$TP),]
completeBloom<-rbind(hadTP, avgTPbloom)

#split by invasion status 
invadedBlooms<-completeBloom[completeBloom$Dreissenids == "Invaded",]
uninvadedBlooms<-completeBloom[completeBloom$Dreissenids == "Uninvaded",]
```

### TP and Microcystin

```{r}
#create linear models 
invaded.lm<-lm(log(invadedBlooms$ESF_Microcystin+.1)~log(invadedBlooms$TP+.1))
uninvaded.lm<-lm(log(uninvadedBlooms$ESF_Microcystin+.1)~log(uninvadedBlooms$TP+.1))

summary(invaded.lm)
summary(uninvaded.lm)

invaded.lm.sum<-summary(invaded.lm)
uninvaded.lm.sum<-summary(uninvaded.lm)
```

```{r}
ggplot(completeBloom, aes(x=TP, y=ESF_Microcystin, color=Dreissenids, shape=Dreissenids)) +
  geom_point(cex=2) + 
  geom_smooth(data=subset(completeBloom, Dreissenids == "Uninvaded" ),
               aes(x=TP, y=ESF_Microcystin, color=Dreissenids), method=lm, se=FALSE)+
  theme_classic()+ 
  scale_x_continuous(trans='log10') +
  scale_y_continuous(trans='log10', breaks=c(1,100, 10000)) +
  scale_color_grey()
  

ggsave("./output_figures/Fig 2-1 Global.png")
```

### Chlorophyll-a and Microcystin 

```{r}
#create linear models
chl.invaded.lm<-lm(log(invadedBlooms$ESF_Microcystin+.1)~log(invadedBlooms$ESF_FP_Chl.a+.1))
chl.uninvaded.lm<-lm(log(uninvadedBlooms$ESF_Microcystin+.1)~log(uninvadedBlooms$ESF_FP_Chl.a+.1))

summary(chl.invaded.lm)
summary(chl.uninvaded.lm)

chl.invaded.lm.sum<-summary(chl.invaded.lm)
chl.uninvaded.lm.sum<-summary(chl.uninvaded.lm)
```

```{r}
ggplot(completeBloom, aes(x=ESF_FP_Chl.a, y=ESF_Microcystin, color=Dreissenids, shape=Dreissenids)) +
  geom_point(cex=2) + 
  geom_smooth(method=lm, se=FALSE)+
  theme_classic()+ 
  scale_x_continuous(trans='log10') +
  scale_y_continuous(trans='log10') + 
  scale_color_grey()
  

ggsave("./output_figures/Fig 2-2 Global.png")
```

## Reduced Dataset
```{r}
#Extracting average annual TP for each lake 
library(plyr)
redavgTP<-ddply(redOWCSLAP, c("LakeID", "Sample_Year"), summarize, 
             Mean = mean(TP, na.rm=TRUE))
```

```{r}
#Taking only rows where there is a value for microcystin
redBloom<-redCSLAP[!is.na(redCSLAP$ESF_Microcystin),]

#Which rows are missing TP values? 
rednoTPbloom<-redBloom[is.na(redBloom$TP),]

#merge average TP to noTPbloom data frame 
redavgTPbloom<-merge(rednoTPbloom, redavgTP, by=c("LakeID", "Sample_Year"), all.x=TRUE, all.y=FALSE)

#remove original TP column, and re-name the average TP column 
redavgTPbloom<-redavgTPbloom[, c(1:16, 18:55)]
colnames(redavgTPbloom)[colnames(redavgTPbloom)=="Mean"] <- "TP"

#re-order columns 
redavgTPbloom<-redavgTPbloom[, c(1:16,54,17:53)]

#combine two data frames
redhadTP<-redBloom[!is.na(redBloom$TP),]
redcompleteBloom<-rbind(redhadTP, redavgTPbloom)

#split by invasion status 
redinvadedBlooms<-redcompleteBloom[redcompleteBloom$Dreissenids == "Invaded",]
reduninvadedBlooms<-redcompleteBloom[redcompleteBloom$Dreissenids == "Uninvaded",]
```


### TP and Microcystin
```{r}
#create linear models 
red.invaded.lm<-lm(log(redinvadedBlooms$ESF_Microcystin+.1)~log(redinvadedBlooms$TP+.1))
red.uninvaded.lm<-lm(log(reduninvadedBlooms$ESF_Microcystin+.1)~log(reduninvadedBlooms$TP+.1))

summary(red.invaded.lm)
summary(red.uninvaded.lm)

red.invaded.lm.sum<-summary(red.invaded.lm)
red.uninvaded.lm.sum<-summary(red.uninvaded.lm)
```

```{r}
ggplot(redcompleteBloom, aes(x=TP, y=ESF_Microcystin, color=Dreissenids, shape=Dreissenids)) +
  geom_point(cex=2) + 
  geom_smooth(data=subset(redcompleteBloom, Dreissenids == "Uninvaded" ),
               aes(x=TP, y=ESF_Microcystin, color=Dreissenids), method=lm, se=FALSE)+
  theme_classic()+ 
  scale_x_continuous(trans='log10') +
  scale_y_continuous(trans='log10', breaks=c(1,100, 10000)) + 
  scale_color_grey()
  

ggsave("./output_data/Fig 2-1 Reduced.png")
```

### Chlorophyll and Microcystin 
```{r}
#create linear models
red.chl.invaded.lm<-lm(log(redinvadedBlooms$ESF_Microcystin+.1)~log(redinvadedBlooms$ESF_FP_Chl.a+.1))
red.chl.uninvaded.lm<-lm(log(reduninvadedBlooms$ESF_Microcystin+.1)~log(reduninvadedBlooms$ESF_FP_Chl+.1))

summary(red.chl.invaded.lm)
summary(red.chl.uninvaded.lm)

red.chl.invaded.lm.sum<-summary(red.chl.invaded.lm)
red.chl.uninvaded.lm.sum<-summary(red.chl.uninvaded.lm)
```

```{r}
ggplot(redcompleteBloom, aes(x=ESF_FP_Chl.a, y=ESF_Microcystin, color=Dreissenids, shape=Dreissenids)) +
  geom_point(cex=2) + 
  geom_smooth(method=lm, se=FALSE)+
  theme_classic()+ 
  scale_x_continuous(trans='log10') +
  scale_y_continuous(trans='log10') + 
  scale_color_grey()
  

ggsave("./output_figures/Fig 2-2 Reduced.png")
```