---
title: "CSLAP MEM March 2020"
author: "Victoria_Field"
date: "2/28/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(car)
library(lme4)
library(ggplot2)
require(lmerTest)
library(lattice)
library(predictmeans)
library(dplyr)
```

#Reading in and organizing datasets
```{r}
CSLAP<-read.csv("CSLAP_Dataset_09232019.csv", header=TRUE)
CSLAP$Sample_Year<-as.factor(CSLAP$Sample_Year)
CSLAP$Sample_Month<-as.factor(CSLAP$Sample_Month)

OWCSLAP<-CSLAP[CSLAP$Info_Type == "OW",]
BSCSLAP<-CSLAP[CSLAP$Info_Type == "BS",]
SBCSLAP<-CSLAP[CSLAP$Info_Type == "SB",]
noSBCSLAP<-CSLAP[CSLAP$Info_Type != "SB",]

df_match<-read.csv("df_match.csv")
```

##Create reduced dataset based on matched lakes and write new CSV
```{r}
redCSLAP<-merge(CSLAP, df_match, by="LakeID")

write.csv(redCSLAP, "redCSLAP.csv")
```

##Bring redCSLAP back in as new data frame to remove unused levels of factor 'LakeID'
```{r}
redCSLAP<-read.csv("redCSLAP.csv")

redOWCSLAP<-redCSLAP[redCSLAP$Info_Type == "OW",]
redBSCSLAP<-redCSLAP[redCSLAP$Info_Type == "BS",]
redSBCSLAP<-redCSLAP[redCSLAP$Info_Type == "SB",]
rednoSBCSLAP<-redCSLAP[redCSLAP$Info_Type != "SB",]
```

#Mixed Effects Models for the **Global Dataset** *(n=69)*

##Open Water Total Phosphorus 
```{r}
one <- OWCSLAP %>% select(TP_mg.L, Dreissenids, CA.SA, Mean_Depth_m, Sample_Year, Sample_Month, LakeID)
one <- one[complete.cases(one), ]

# Fit the model
fit <- lmer(log(TP_mg.L+.01) ~ Dreissenids + log(CA.SA) + Mean_Depth_m + (1|Sample_Year) + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=one)

# Obtain predicted and residual values
one$predicted <- predict(fit)
one$residuals <- residuals(fit)

head(one)
```

```{r}
OWTP<-lmer(log(TP_mg.L+.01) ~ Dreissenids + log(CA.SA) + Mean_Depth_m + (1|Sample_Year) + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=one)

summary(OWTP)
```

```{r}
plot(OWTP,col=one$LakeID)
plot(OWTP,col=one$LakeID, id=0.5, idLabels=~.obs, cex=.5)

plot(OWTP,col=one$Sample_Year)
plot(OWTP,col=one$Sample_Year, id=0.5, idLabels=~.obs, cex=.5)

plot(OWTP,LakeID~resid(.))

qqmath(OWTP, col=one$LakeID, key=list(text=list(levels(one$LakeID))))

qqnorm(resid(OWTP))
qqline(resid(OWTP))
```

```{r}
# Select out data of interest:
d <- OWCSLAP %>% select(TP_mg.L, Dreissenid, wt, disp)

# Fit the model
fit <- lm(mpg ~ hp + wt+ disp, data = d)

# Obtain predicted and residual values
d$predicted <- predict(fit)
d$residuals <- residuals(fit)
```

##Bottom Sample Total Phosphorus 
-**Updated 03/08** `boundary (singular) fit` removed random effect of Sample_Month (variance =0.00), but then output `Model failed to converge` warning 

-**Updated 03/09** `Model failed to converge` fixed by taking log of mean_depth_m and adding only .01 to `log(TP_mg.L)` instead of 1
```{r}
BSTP<-lmer(log(TP_mg.L+.01) ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m)  + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=BSCSLAP)

summary(BSTP)
```

```{r}
plot(BSTP)

qqnorm(resid(BSTP))
qqline(resid(BSTP))
```

##Open Water Chlorophyll a
```{r}
Chl<-lmer(log(Extracted_Chl.a_ug.L+1) ~ Dreissenids + log(CA.SA) + Mean_Depth_m + (1|Sample_Year) + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=OWCSLAP)

summary(Chl)
```

```{r}
#residual plot
plot(Chl)

#fitted versus observed

##remove NAs from original variable vector so it can be plotted against the fitted values from the model
Chl2<-OWCSLAP$Extracted_Chl.a_ug.L[!is.na(OWCSLAP$Extracted_Chl.a_ug.L)]
plot(fitted(Chl), log(Chl2))
abline(a=0,b=1)

qqnorm(resid(Chl))
qqline(resid(Chl))
```

##Open Water True Color
```{r}
OWTC<-lmer(log(True_Color_PTU) ~ Dreissenids + log(CA.SA) + Mean_Depth_m + (1|Sample_Year) + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=OWCSLAP)

summary(OWTC)
```

```{r}
plot(OWTC)

##remove NAs from original variable vector so it can be plotted against the fitted values from the model
OWTC2<-OWCSLAP$True_Color_PTU[!is.na(OWCSLAP$True_Color_PTU)]
plot(fitted(OWTC), log(OWTC2))
abline(a=0,b=1)

qqnorm(resid(OWTC))
qqline(resid(OWTC))
```

##Open Water Total Nitrogen
```{r}
OWTN<-lmer(log(TN_mg.L+1) ~ Dreissenids + log(CA.SA) + Mean_Depth_m + (1|Sample_Year) + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=OWCSLAP)

summary(OWTN)
```

```{r}
plot(OWTN)

qqnorm(resid(OWTN))
qqline(resid(OWTN))
```

##Secchi Depth
-**Update 03/08** `Model failed to converge` removed log-transformation of Secchi to resolve
```{r}
Secchi<-lmer(Secchi_Depth_m ~ Dreissenids + log(CA.SA) + Mean_Depth_m + (1|Sample_Year) + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=noSBCSLAP)

summary(Secchi)
```

```{r}
plot(Secchi)

##remove NAs from original variable vector so it can be plotted against the fitted values from the model
Secchi2<-noSBCSLAP$Secchi_Depth_m[!is.na(noSBCSLAP$Secchi_Depth_m)]
plot(fitted(Secchi), Secchi2)
abline(a=0,b=1)

qqnorm(resid(Secchi))
qqline(resid(Secchi))
```


##Shoreline Bloom Chlorophyll a 
-**update 03/08** `boundary (singular) fit` Removed random effect of Sample_Year with ~0.00 variance to resolve warning.
```{r}
SBChl<-lmer(log(ESF_Chl.a_ug.L) ~ Dreissenids + log(CA.SA) + Mean_Depth_m  + (1|Sample_Month) + (1|LakeID) , data=SBCSLAP)

summary(SBChl)
```

```{r}
plot(SBChl)

##remove NAs from original variable vector so it can be plotted against the fitted values from the model
SBChl2<-SBCSLAP$ESF_Chl.a_ug.L[!is.na(SBCSLAP$ESF_Chl.a_ug.L)]
plot(fitted(SBChl), log(SBChl2))
abline(a=0,b=1)

qqnorm(resid(SBChl))
qqline(resid(SBChl))
```

##Shoreline Bloom Microcystin
 -**update 03/08** `boundary (singular) fit` Removed random effect of Sample_Month with 0.00 variance to resolve warning.
```{r}
SBmicro<-lmer(log(ESF_Microcystin_ug.L) ~ Dreissenids + log(CA.SA) + Mean_Depth_m + (1|Sample_Year)  + (1|LakeID) + (1|Sample_Year:LakeID), data=SBCSLAP)

summary(SBmicro)
```

```{r}
plot(SBmicro)

##remove NAs from original variable vector so it can be plotted against the fitted values from the model
SBmicro2<-SBCSLAP$ESF_Microcystin_ug.L[!is.na(SBCSLAP$ESF_Microcystin_ug.L)]
plot(fitted(SBmicro), log(SBmicro2))
abline(a=0,b=1)

qqnorm(resid(SBmicro))
qqline(resid(SBmicro))
```

#Mixed Effects Models for the **Reduced Dataset** *(n=16)*

##Open Water Total Phosphorus 
  -**update 03/08** `Model failed to converge.` Removed random effect of LakeID to ~0.00 variance to resolve warning.
```{r}
redOWTP<-lmer(log(TP_mg.L+1) ~ Dreissenids.x  + (1|Sample_Year) + (1|Sample_Month)  + (1|Sample_Year:LakeID), data=redOWCSLAP)

summary(redOWTP)
```

```{r}
plot(redOWTP)

qqnorm(resid(redOWTP))
qqline(resid(redOWTP))
```

##Bottom Sample Total Phosphorus 
-**Update 03/08** `boundary (singular) fit` removed random effect of LakeID (variance = 0.00) to resolve warning
```{r}
redBSTP<-lmer(log(TP_mg.L+1) ~ Dreissenids.x + (1|Sample_Year) + (1|Sample_Month)  + (1|Sample_Year:LakeID) + (1|Sample_Month:LakeID), data=redBSCSLAP)

summary(redBSTP)
```

```{r}
plot(redBSTP)

qqnorm(resid(redBSTP))
qqline(resid(redBSTP))
```


##Open Water Chlorophyll a
-**Update 03/10** `Model failed to converge` fixed by adding .01 to `log(Extracted_Chl.a_ug.L)` instead of 1 
```{r}
redChl<-lmer(log(Extracted_Chl.a_ug.L+.01) ~ Dreissenids.x  + (1|Sample_Year) + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=redOWCSLAP)

summary(redChl)
```

```{r}
plot(redChl)

##remove NAs from original variable vector so it can be plotted against the fitted values from the model
redChl2<-redOWCSLAP$Extracted_Chl.a_ug.L[!is.na(redOWCSLAP$Extracted_Chl.a_ug.L)]
plot(fitted(redChl), log(redChl2))
abline(a=0,b=1)

qqnorm(resid(redChl))
qqline(resid(redChl))
```

##Open Water True Color
**Update 03/31** Removed `Sample_Month:LakeID` to alleviate Convergence Code error
```{r}
redOWTC<-lmer(log(True_Color_PTU) ~ Dreissenids.x  + (1|Sample_Year) + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=redOWCSLAP)

summary(redOWTC)
```

```{r}
plot(redOWTC)

##remove NAs from original variable vector so it can be plotted against the fitted values from the model
redOWTC2<-redOWCSLAP$True_Color_PTU[!is.na(redOWCSLAP$True_Color_PTU)]
plot(fitted(redOWTC), log(redOWTC2))
abline(a=0,b=1)

qqnorm(resid(redOWTC))
qqline(resid(redOWTC))
```

##Open Water Total Nitrogen
-**Update 03/10** `Model failed to converge` resolved by removing low variance factor of `Sample_Month` and adding .01 to `log(TN_mg.L)` instead of 1.0 
```{r}
redOWTN<-lmer(log(TN_mg.L+.01) ~ Dreissenids.x  + (1|Sample_Year)  + (1|LakeID) + (1|Sample_Year:LakeID) + (1|Sample_Month:LakeID), data=redOWCSLAP)

summary(redOWTN)
```

```{r}
plot(redOWTN)

redOWTN2<-redOWCSLAP$TN_mg.L[!is.na(redOWCSLAP$TN_mg.L)]
plot(fitted(redOWTN), log(redOWTN2))
abline(a=0,b=1)

qqnorm(resid(redOWTN))
qqline(resid(redOWTN))
```

##Secchi Depth
```{r}
redSecchi<-lmer(log(Secchi_Depth_m) ~ Dreissenids.x + (1|Sample_Year) + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID) + (1|Sample_Month:LakeID), data=rednoSBCSLAP)

summary(redSecchi)
```

```{r}
plot(redSecchi)

redSecchi2<-rednoSBCSLAP$Secchi_Depth_m[!is.na(rednoSBCSLAP$Secchi_Depth_m)]
plot(fitted(redSecchi), log(redSecchi2))
abline(a=0,b=1)

qqnorm(resid(redSecchi))
qqline(resid(redSecchi))
```

##Shoreline Bloom Chlorophyll a 
-**update 03/08** `boundary (singular) fit` removed random effect of Sample_Year and Sample_Year:LakeID to resolve
```{r}
redSBChl<-lmer(log(ESF_Chl.a_ug.L) ~ Dreissenids.x  + (1|Sample_Month) + (1|LakeID), data=redSBCSLAP)

summary(redSBChl)
```

```{r}
plot(redSBChl)

redSBChl2<-redSBCSLAP$ESF_Chl.a_ug.L[!is.na(redSBCSLAP$ESF_Chl.a_ug.L)]
plot(fitted(redSBChl), log(redSBChl2))
abline(a=0,b=1)

qqnorm(resid(redSBChl))
qqline(resid(redSBChl))
```

##Shoreline Bloom Microcystin
-**Update 03/08** `boundary (singular) fit`, `model matrix error`, and `model failed to converge` so removed random effect of Sample_Year:LakeID, random effect of month, and fixed effect of Mean_Depth_m to resolve
```{r}
redMicro<-lmer(log(ESF_Microcystin_ug.L+1) ~ Dreissenids.x + (1|Sample_Year)  + (1|LakeID), data=redSBCSLAP)

summary(redMicro)
```

```{r}
plot(redMicro)

redMicro2<-redSBCSLAP$ESF_Microcystin_ug.L[!is.na(redSBCSLAP$ESF_Microcystin_ug.L)]
plot(fitted(redMicro), log(redMicro2))
abline(a=0,b=1)

qqnorm(resid(redMicro))
qqline(resid(redMicro))
```

