---
title: "CSLAP Gamma Models May 2020"
author: "Victoria Field"
date: "5/15/2020"
output: 
  html_document:
    toc: true
    toc_float: true
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(car)
library(lme4)
library(ggplot2)
require(lmerTest)
library(lattice)
library(predictmeans)
library(dplyr)
library(tidyr)
library(DHARMa)
library(magrittr)
library(pryr)
library(MASS)
```

#Data Read In 
```{r}
CSLAP<-read.csv("CSLAP_Dataset_09232019.csv", header=TRUE)
CSLAP$Sample_Year<-as.factor(CSLAP$Sample_Year)
CSLAP$Sample_Month<-as.factor(CSLAP$Sample_Month)

#Create TN:TP column 
CSLAP$TN_TP<-(CSLAP$TN_mg.L+.01)/(CSLAP$TP_mg.L+.01)

#Split by `Info_Type`
OWCSLAP<-CSLAP[CSLAP$Info_Type == "OW",]
BSCSLAP<-CSLAP[CSLAP$Info_Type == "BS",]
SBCSLAP<-CSLAP[CSLAP$Info_Type == "SB",]
noSBCSLAP<-CSLAP[CSLAP$Info_Type != "SB",]

redCSLAP<-read.csv("redCSLAP.csv")

#Create TN:TP column 
redCSLAP$TN_TP<-(redCSLAP$TN_mg.L+.01)/(redCSLAP$TP_mg.L+.01)

redOWCSLAP<-redCSLAP[redCSLAP$Info_Type == "OW",]
redBSCSLAP<-redCSLAP[redCSLAP$Info_Type == "BS",]
redSBCSLAP<-redCSLAP[redCSLAP$Info_Type == "SB",]
rednoSBCSLAP<-redCSLAP[redCSLAP$Info_Type != "SB",]
```

#Gamma(link="identity") models for the **Global Dataset** *(n=69

##Open Water TN:TP

```{r}
TNTP<-glm(TN_TP ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m), data=OWCSLAP, family=Gamma(link="identity"))

summary(TNTP)

plot(TNTP)

qqPlot(resid(TNTP))

simTNTP  <-  simulateResiduals(TNTP)
plotSimulatedResiduals(simTNTP) 
```

##Open Water Total Phosphorus 

###Remove troublesome `LakeID` and `Sample_Year`
```{r}
#Select out data of interest
df <- OWCSLAP %>% dplyr::select(TP_mg.L, Dreissenids, CA.SA, Mean_Depth_m, Sample_Year, Sample_Month, LakeID)
df <- df[complete.cases(df), ]

#Remove lakes and year 2015
df <- df[df$LakeID != "0601GUI0188",]
df <- df[df$LakeID != "1201ECA0697", ]
df <- df[df$LakeID != "1701LLO0708", ]
df <- df[df$LakeID != "1201GAL0563", ]
df <- df[df$Sample_Year != "2015",]
df$LakeID <- as.factor(as.character(df$LakeID))
```

```{r}
OWTP<-glm(TP_mg.L+1 ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m), data=df, family=Gamma(link="identity"))

summary(OWTP)

plot(OWTP)

qqPlot(resid(OWTP))

simOWTP  <-  simulateResiduals(OWTP)
plotSimulatedResiduals(simOWTP) 
```

##Bottom Sample TP 

```{r}
BSTP<-glm(TP_mg.L+1 ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m), data=BSCSLAP, family=Gamma(link="identity"))

summary(BSTP)

plot(BSTP)

qqPlot(resid(BSTP))

simBSTP  <-  simulateResiduals(BSTP)
plotSimulatedResiduals(simBSTP) 
```

##Open Water TN 

```{r}
OWTN<-glm(TN_mg.L+1 ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m), data=OWCSLAP, family=Gamma(link="identity"))

summary(OWTN)

plot(OWTN)

qqPlot(resid(OWTN))

simOWTN  <-  simulateResiduals(OWTN)
plotSimulatedResiduals(simOWTN) 
```

##Open Water Chlorophyll-a 

```{r}
OWChl<-glm(Extracted_Chl.a_ug.L+.01 ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m), data=OWCSLAP, family=Gamma(link="identity"))

summary(OWChl)

plot(OWChl)

qqPlot(resid(OWChl))

simOWChl  <-  simulateResiduals(OWChl)
plotSimulatedResiduals(simOWChl) 
```

##Secchi Depth

```{r}
OWSecchi<-glm(Secchi_Depth_m ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m), data=OWCSLAP, family=Gamma(link="identity"))

summary(OWSecchi)

plot(OWSecchi)

qqPlot(resid(OWSecchi))

simOWSecchi  <-  simulateResiduals(OWSecchi)
plotSimulatedResiduals(simOWSecchi) 
```

##True Color

```{r}
OWTC<-glm(True_Color_PTU ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m), data=OWCSLAP, family=Gamma(link="identity"))

summary(OWTC)

plot(OWTC)

qqPlot(resid(OWTC))

simOWTC  <-  simulateResiduals(OWTC)
plotSimulatedResiduals(simOWTC) 
```

##Shoreline Bloom Chlorophyll

```{r}
SBChl<-glm(ESF_Chl.a_ug.L+1 ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m), data=SBCSLAP, family=Gamma(link="identity"))

summary(SBChl)

plot(SBChl)

qqPlot(resid(SBChl))

simSBChl  <-  simulateResiduals(SBChl)
plotSimulatedResiduals(simSBChl) 
```

##Shoreline Bloom Microcystin

```{r}
SBmicro<-glm(ESF_Microcystin_ug.L+1 ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m), data=SBCSLAP, family=Gamma(link="identity"))

summary(SBmicro)

plot(SBmicro)

qqPlot(resid(SBmicro))

simSBmicro  <-  simulateResiduals(SBmicro)
plotSimulatedResiduals(simSBmicro) 
```

#Gamma(link="log") models for the **Global Dataset** *(n=69)*

##Open Water TN:TP

```{r}
TNTP2<-glm(TN_TP ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m), data=OWCSLAP, family=Gamma(link="log"))

summary(TNTP2)

plot(TNTP2)

qqPlot(resid(TNTP2))

simTNTP2  <-  simulateResiduals(TNTP2)
plotSimulatedResiduals(simTNTP2) 
```

##Open Water Total Phosphorus 

```{r}
OWTP2<-glm(TP_mg.L+1 ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m), data=df, family=Gamma(link="log"))

summary(OWTP2)

plot(OWTP2)

qqPlot(resid(OWTP2))

simOWTP2  <-  simulateResiduals(OWTP2)
plotSimulatedResiduals(simOWTP2) 
```

##Bottom Sample TP 

```{r}
BSTP2<-glm(TP_mg.L+1 ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m), data=BSCSLAP, family=Gamma(link="log"))

summary(BSTP2)

plot(BSTP2)

qqPlot(resid(BSTP2))

simBSTP2  <-  simulateResiduals(BSTP2)
plotSimulatedResiduals(simBSTP2) 
```

##Open Water TN 

```{r}
OWTN2<-glm(TN_mg.L+1 ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m), data=OWCSLAP, family=Gamma(link="log"))

summary(OWTN2)

plot(OWTN2)

qqPlot(resid(OWTN2))

simOWTN2  <-  simulateResiduals(OWTN2)
plotSimulatedResiduals(simOWTN2) 
```

##Open Water Chlorophyll-a 

```{r}
OWChl2<-glm(Extracted_Chl.a_ug.L+.01 ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m), data=OWCSLAP, family=Gamma(link="log"))

summary(OWChl2)

plot(OWChl2)

qqPlot(resid(OWChl2))

simOWChl2  <-  simulateResiduals(OWChl2)
plotSimulatedResiduals(simOWChl2) 
```

##Secchi Depth

```{r}
OWSecchi2<-glm(Secchi_Depth_m ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m), data=OWCSLAP, family=Gamma(link="log"))

summary(OWSecchi2)

plot(OWSecchi2)

qqPlot(resid(OWSecchi2))

simOWSecchi2  <-  simulateResiduals(OWSecchi2)
plotSimulatedResiduals(simOWSecchi2) 
```

##True Color

```{r}
OWTC2<-glm(True_Color_PTU ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m), data=OWCSLAP, family=Gamma(link="log"))

summary(OWTC2)

plot(OWTC2)

qqPlot(resid(OWTC2))

simOWTC2  <-  simulateResiduals(OWTC2)
plotSimulatedResiduals(simOWTC2) 
```

##Shoreline Bloom Chlorophyll

```{r}
SBChl2<-glm(ESF_Chl.a_ug.L+1 ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m), data=SBCSLAP, family=Gamma(link="log"))

summary(SBChl2)

plot(SBChl2)

qqPlot(resid(SBChl2))

simSBChl2  <-  simulateResiduals(SBChl2)
plotSimulatedResiduals(simSBChl2) 
```

##Shoreline Bloom Microcystin

```{r}
SBmicro2<-glm(ESF_Microcystin_ug.L+1 ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m), data=SBCSLAP, family=Gamma(link="log"))

summary(SBmicro2)

plot(SBmicro2)

qqPlot(resid(SBmicro2))

simSBmicro2  <-  simulateResiduals(SBmicro2)
plotSimulatedResiduals(simSBmicro2) 
```

#Gamma(link="inverse") models for the **Global Dataset** *(n=69)*

##Open Water Total Phosphorus 

```{r}
OWTP3<-glm(TP_mg.L+1 ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m), data=df, family=Gamma(link="inverse"))

summary(OWTP3)

plot(OWTP3)

qqPlot(resid(OWTP3))

simOWTP3  <-  simulateResiduals(OWTP3)
plotSimulatedResiduals(simOWTP3) 
```

##Bottom Sample TP 

```{r}
BSTP3<-glm(TP_mg.L+1 ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m), data=BSCSLAP, family=Gamma(link="inverse"))

summary(BSTP3)

plot(BSTP3)

qqPlot(resid(BSTP3))

simBSTP3  <-  simulateResiduals(BSTP3)
plotSimulatedResiduals(simBSTP3) 
```

##Open Water TN 

```{r}
OWTN3<-glm(TN_mg.L+1 ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m), data=OWCSLAP, family=Gamma(link="inverse"))

summary(OWTN3)

plot(OWTN3)

qqPlot(resid(OWTN3))

simOWTN3  <-  simulateResiduals(OWTN3)
plotSimulatedResiduals(simOWTN3) 
```

##Open Water Chlorophyll-a 

```{r}
OWChl3<-glm(Extracted_Chl.a_ug.L+.01 ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m), data=OWCSLAP, family=Gamma(link="inverse"))

summary(OWChl3)

plot(OWChl3)

qqPlot(resid(OWChl3))

simOWChl3  <-  simulateResiduals(OWChl3)
plotSimulatedResiduals(simOWChl3) 
```

##Secchi Depth

```{r}
OWSecchi3<-glm(Secchi_Depth_m ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m), data=OWCSLAP, family=Gamma(link="inverse"))

summary(OWSecchi3)

plot(OWSecchi3)

qqPlot(resid(OWSecchi3))

simOWSecchi3  <-  simulateResiduals(OWSecchi3)
plotSimulatedResiduals(simOWSecchi3) 
```

##True Color

```{r}
OWTC3<-glm(True_Color_PTU ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m), data=OWCSLAP, family=Gamma(link="inverse"))

summary(OWTC3)

plot(OWTC3)

qqPlot(resid(OWTC3))

simOWTC3  <-  simulateResiduals(OWTC3)
plotSimulatedResiduals(simOWTC3) 
```

##Shoreline Bloom Chlorophyll
`**Model won't run**
```{r, eval=FALSE}
SBChl3<-glm(ESF_Chl.a_ug.L+.01 ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m), data=SBCSLAP, family=Gamma(link="inverse"))

summary(SBChl3)

plot(SBChl3)

qqPlot(resid(SBChl3))

simSBChl3  <-  simulateResiduals(SBChl3)
plotSimulatedResiduals(simSBChl3) 
```

##Shoreline Bloom Microcystin
`**Model won't run**
```{r, eval=FALSE}
SBmicro3<-glm(ESF_Microcystin_ug.L+.01 ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m), data=SBCSLAP, family=Gamma(link="inverse"))

summary(SBmicro3)

plot(SBmicro3)

qqPlot(resid(SBmicro3))

simSBmicro3  <-  simulateResiduals(SBmicro3)
plotSimulatedResiduals(simSBmicro3) 
```

#Link function comparisons by variable. Order = "Identity", "Log", "Inverse"

##Open Water TP 
```{r}
plot(OWTP)
plot(OWTP2)
plot(OWTP3)
```

##Bottom Sample TP
```{r}
plot(BSTP)
plot(BSTP2)
plot(BSTP3)
```

##Open Water TN
```{r}
plot(OWTN)
plot(OWTN2)
plot(OWTN3)
```

##Open Water Chlorophyll-a
```{r}
plot(OWChl)
plot(OWChl2)
plot(OWChl3)
```

##Open Water Secchi depth
```{r}
plot(OWSecchi)
plot(OWSecchi2)
plot(OWSecchi3)
```

##Open Water True Color
```{r}
plot(OWTC)
plot(OWTC2)
plot(OWTC3)
```

##Shoreline Bloom Chlorophyll 
-**NOTE** Gamma("inverse") did not work
```{r}
plot(SBChl)
plot(SBChl2)
```

##Shoreline Bloom Microcystin 
-**NOTE** Gamma("inverse") did not work
```{r}
plot(SBmicro)
plot(SBmicro2)
```

