---
title: "CSLAP MEMs May 2020"
author: "Victoria Field"
date: "5/15/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(car)
library(lme4)
library(ggplot2)
require(lmerTest)
library(lattice)
library(predictmeans)
library(dplyr)
library(tidyr)
library(DHARMa)
library(magrittr)
library(pryr)
library(MASS)
```

#Data Read In 
```{r}
CSLAP<-read.csv("CSLAP_Dataset_09232019.csv", header=TRUE)
CSLAP$Sample_Year<-as.factor(CSLAP$Sample_Year)
CSLAP$Sample_Month<-as.factor(CSLAP$Sample_Month)

#Create TN:TP column 
CSLAP$TP_TN<-CSLAP$TN_mg.L/CSLAP$TP_mg.L

#Split by `Info_Type`
OWCSLAP<-CSLAP[CSLAP$Info_Type == "OW",]
BSCSLAP<-CSLAP[CSLAP$Info_Type == "BS",]
SBCSLAP<-CSLAP[CSLAP$Info_Type == "SB",]
noSBCSLAP<-CSLAP[CSLAP$Info_Type != "SB",]


redCSLAP<-read.csv("redCSLAP.csv")

#Create TN:TP column 
redCSLAP$TP_TN<-redCSLAP$TN_mg.L/redCSLAP$TP_mg.L

redOWCSLAP<-redCSLAP[redCSLAP$Info_Type == "OW",]
redBSCSLAP<-redCSLAP[redCSLAP$Info_Type == "BS",]
redSBCSLAP<-redCSLAP[redCSLAP$Info_Type == "SB",]
rednoSBCSLAP<-redCSLAP[redCSLAP$Info_Type != "SB",]
```

#Mixed Effects Models for the **Global Dataset** *(n=69)*

##Open Water Total Phosphorus 

###Remove troublesome `LakeID` and `Sample_Year`
```{r}
#Select out data of interest
df <- OWCSLAP %>% dplyr::select(TP_mg.L, Dreissenids, CA.SA, Mean_Depth_m, Sample_Year, Sample_Month, LakeID)
df <- df[complete.cases(df), ]

#Remove lakes and year 2015
df <- df[df$LakeID != "0601GUI0188",]
df <- df[df$LakeID != "1201ECA0697", ]
df <- df[df$LakeID != "1701LLO0708", ]
df <- df[df$LakeID != "1201GAL0563", ]
df <- df[df$Sample_Year != "2015",]
df$LakeID <- as.factor(as.character(df$LakeID))
```

###Fit the model
```{r}
OWTP<-lmer(log(TP_mg.L+.01) ~ Dreissenids + log(CA.SA) + Mean_Depth_m + (1|Sample_Year) + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=df)

summary(OWTP)
```

###Model diagnostics
```{r}
plot(OWTP)

qqmath(OWTP)

qqPlot(resid(OWTP))
```

##Bottom Sample Total Phosphorus 
-**Updated 03/08** `boundary (singular) fit` removed random effect of Sample_Month (variance =0.00), but then output `Model failed to converge` warning 

-**Updated 03/09** `Model failed to converge` fixed by taking log of mean_depth_m and adding only .01 to `log(TP_mg.L)` instead of 1

###Fit the model
```{r}
BSTP<-lmer(log(TP_mg.L+.01) ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m)  + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=BSCSLAP)

summary(BSTP)
```

###Model diagnostics
```{r}
plot(BSTP)

qqPlot(resid(BSTP))
```

##Open Water Chlorophyll a

###Fit the model
```{r}
Chl<-lmer(log(Extracted_Chl.a_ug.L+1) ~ Dreissenids + log(CA.SA) + Mean_Depth_m + (1|Sample_Year) + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=OWCSLAP)

summary(Chl)
```

###Model diagnostics
```{r}
plot(Chl)

qqPlot(resid(Chl))
```

##Open Water True Color

###Fit the model
```{r}
OWTC<-lmer(log(True_Color_PTU) ~ Dreissenids + log(CA.SA) + Mean_Depth_m + (1|Sample_Year) + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=OWCSLAP)

summary(OWTC)
```

###Model diagnostics
```{r}
plot(OWTC)

qqPlot(resid(OWTC))
```

##Open Water Total Nitrogen

###Fit the model
```{r}
OWTN<-lmer(log(TN_mg.L+1) ~ Dreissenids + log(CA.SA) + Mean_Depth_m + (1|Sample_Year) + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=OWCSLAP)

summary(OWTN)
```

```{r}
plot(OWTN)

qqPlot(resid(OWTN))
```

##Secchi Depth
-**Update 03/08** `Model failed to converge` removed log-transformation of Secchi to resolve

###Fit the model
```{r}
Secchi<-lmer(Secchi_Depth_m ~ Dreissenids + log(CA.SA) + Mean_Depth_m + (1|Sample_Year) + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=noSBCSLAP)

summary(Secchi)
```

###Model Diagnostics
```{r}
plot(Secchi)

qqPlot(resid(Secchi))
```


##Shoreline Bloom Chlorophyll a 
-**update 03/08** `boundary (singular) fit` Removed random effect of Sample_Year with ~0.00 variance to resolve warning.

###Fit the model
```{r}
SBChl<-lmer(log(ESF_Chl.a_ug.L) ~ Dreissenids + log(CA.SA) + Mean_Depth_m  + (1|Sample_Month) + (1|LakeID) , data=SBCSLAP)

summary(SBChl)
```

###Model diagnostics
```{r}
plot(SBChl)

qqPlot(resid(SBChl))
```

##Shoreline Bloom Microcystin
 -**update 03/08** `boundary (singular) fit` Removed random effect of Sample_Month with 0.00 variance to resolve warning.
 
###Fit the model
```{r}
SBmicro<-lmer(log(ESF_Microcystin_ug.L) ~ Dreissenids + log(CA.SA) + Mean_Depth_m + (1|Sample_Year)  + (1|LakeID) + (1|Sample_Year:LakeID), data=SBCSLAP)

summary(SBmicro)
```

```{r}
plot(SBmicro)

qqPlot(resid(SBmicro))
```

#Mixed Effects Models for the **Reduced Dataset** *(n=16)*

##Open Water Total Phosphorus 
  -**update 03/08** `Model failed to converge.` Removed random effect of LakeID to ~0.00 variance to resolve warning.

###Fit the model
```{r}
redOWTP<-lmer(log(TP_mg.L+1) ~ Dreissenids.x  + (1|Sample_Year) + (1|Sample_Month)  + (1|Sample_Year:LakeID), data=redOWCSLAP)

summary(redOWTP)
```

###Model Diagnostics
```{r}
plot(redOWTP)

qqPlot(resid(redOWTP))
```

##Bottom Sample Total Phosphorus 
-**Update 03/08** `boundary (singular) fit` removed random effect of LakeID (variance = 0.00) to resolve warning
-**Update 05/15*** `boundary (singular) fit` removed random effect of Sample Year to resolve warning

###Fit the model
```{r}
redBSTP<-lmer(log(TP_mg.L+1) ~ Dreissenids.x  + (1|Sample_Month)  + (1|Sample_Year:LakeID), data=redBSCSLAP)

summary(redBSTP)
```

###Model diagnostics
```{r}
plot(redBSTP)

qqPlot(resid(redBSTP))
```


##Open Water Chlorophyll a
-**Update 03/10** `Model failed to converge` fixed by adding .01 to `log(Extracted_Chl.a_ug.L)` instead of 1 

###Fit the model
```{r}
redChl<-lmer(log(Extracted_Chl.a_ug.L+.01) ~ Dreissenids.x  + (1|Sample_Year) + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=redOWCSLAP)

summary(redChl)
```

###Model diagnostics
```{r}
plot(redChl)

qqPlot(resid(redChl))
```

##Open Water True Color
**Update 03/31** Removed `Sample_Month:LakeID` to alleviate Convergence Code error

###Fit the model
```{r}
redOWTC<-lmer(log(True_Color_PTU) ~ Dreissenids.x  + (1|Sample_Year) + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=redOWCSLAP)

summary(redOWTC)
```

###Model diagnostics
```{r}
plot(redOWTC)

qqPlot(resid(redOWTC))
```

##Open Water Total Nitrogen
-**Update 03/10** `Model failed to converge` resolved by removing low variance factor of `Sample_Month` and adding .01 to `log(TN_mg.L)` instead of 1.0 
-**Update 05/15** `Model failed to converge` resolved by adding 1.0 to `log(TN_mg.L)` instead of 0.1

###Fit the model
```{r}
redOWTN<-lmer(log(TN_mg.L+1) ~ Dreissenids.x  + (1|Sample_Year)  + (1|LakeID) + (1|Sample_Year:LakeID), data=redOWCSLAP)

summary(redOWTN)
```

###Model diagnostics
```{r}
plot(redOWTN)

qqPlot(resid(redOWTN))
```

##Secchi Depth
**Update 05/15** `boundary (singular) fit` resolved by removing `Sample Month` (variance = 0.00)
**Update 05/15** `Model failed to converge` resolved by NOT log transforming `Secchi depth`

###Fit the Model
```{r}
redSecchi<-lmer(Secchi_Depth_m ~ Dreissenids.x + (1|Sample_Year) + (1|LakeID) + (1|Sample_Year:LakeID), data=rednoSBCSLAP)

summary(redSecchi)
```

###Model diagnostics
```{r}
plot(redSecchi)

qqPlot(resid(redSecchi))
```

##Shoreline Bloom Chlorophyll a 
-**update 03/08** `boundary (singular) fit` removed random effect of Sample_Year and Sample_Year:LakeID to resolve

###Fit the model
```{r}
redSBChl<-lmer(log(ESF_Chl.a_ug.L) ~ Dreissenids.x  + (1|Sample_Month) + (1|LakeID), data=redSBCSLAP)

summary(redSBChl)
```

###Model diagnostics
```{r}
plot(redSBChl)

qqPlot(resid(redSBChl))
```

##Shoreline Bloom Microcystin
-**Update 03/08** `boundary (singular) fit`, `model matrix error`, and `model failed to converge` so removed random effect of Sample_Year:LakeID, random effect of month, and fixed effect of Mean_Depth_m to resolve

###Fit the model
```{r}
redMicro<-lmer(log(ESF_Microcystin_ug.L+1) ~ Dreissenids.x + (1|Sample_Year)  + (1|LakeID), data=redSBCSLAP)

summary(redMicro)
```

###Model diagnostics
```{r}
plot(redMicro)

qqPlot(resid(redMicro))
```

