---
title: "CSLAP MEM Diagnostics May 2020"
author: "Victoria Field"
date: "5/12/2020"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(car)
library(lme4)
library(ggplot2)
require(lmerTest)
library(lattice)
library(predictmeans)
library(dplyr)
library(tidyr)
library(DHARMa)
library(magrittr)
library(pryr)
library(MASS)
```

#Data Read In 
```{r}
CSLAP<-read.csv("CSLAP_Dataset_09232019.csv", header=TRUE)
CSLAP$Sample_Year<-as.factor(CSLAP$Sample_Year)
CSLAP$Sample_Month<-as.factor(CSLAP$Sample_Month)

#Split by `Info_Type`
OWCSLAP<-CSLAP[CSLAP$Info_Type == "OW",]
BSCSLAP<-CSLAP[CSLAP$Info_Type == "BS",]
SBCSLAP<-CSLAP[CSLAP$Info_Type == "SB",]
noSBCSLAP<-CSLAP[CSLAP$Info_Type != "SB",]

redCSLAP<-read.csv("redCSLAP.csv")

redOWCSLAP<-redCSLAP[redCSLAP$Info_Type == "OW",]
redBSCSLAP<-redCSLAP[redCSLAP$Info_Type == "BS",]
redSBCSLAP<-redCSLAP[redCSLAP$Info_Type == "SB",]
rednoSBCSLAP<-redCSLAP[redCSLAP$Info_Type != "SB",]
```

#Mixed Effects Models for the **Global Dataset** *(n=69 lakes)*

##Open Water Total Phosphorus 
```{r}
getwd()
#Select out data of interest
df <- OWCSLAP %>% dplyr::select(TP_mg.L, Dreissenids, CA.SA, Mean_Depth_m, Sample_Year, Sample_Month, LakeID)
df <- df[complete.cases(df), ]

#Check distribution 
##gamma
gamma <- fitdistr(df$TP_mg.L+0.01, "gamma")
qqp(df$TP_mg.L+0.01, "gamma", shape = gamma$estimate[[1]], rate = gamma$estimate[[2]])


##log-normal 
qqp(df$TP_mg.L, "lnorm")
```

###Model 1 - Full Model, TP log transformed
```{r}
fit1<-lmer(log(TP_mg.L+.01) ~ Dreissenids + log(CA.SA) + Mean_Depth_m + (1|Sample_Year) + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=df)

summary(fit1)

plot(fit1,col=df$LakeID) #plots model residuals grouped by lake
plot(fit1,col=df$LakeID, id=0.5, idLabels=~.obs, cex=.5) #plots model residuals with labels (observation number) for values outside Normal confidence limits 

plot(fit1,col=df$Sample_Year) #plots model residuals grouped by year 
plot(fit1,col=df$Sample_Year, id=0.5, idLabels=~.obs, cex=.5) #plots model residuals with labels (observation number) for values outside Normal confidence limits 

qqnorm(resid(fit1))
qqline(resid(fit1))
```

From observation numbers, we can see that those points on the far right of each plot correspond to the following lakes and years: Guilford Lake, East Caroga, Little Long Pond, and Galway Lake. And Year 2015 (blue), Year 2012 (black), Year 2017 (Pink), and Year 2016 (Cyan)

###Model 2 - removing "troublesome" lakes 
```{r}
#Select out data of interest
df2 <- df[df$LakeID != "0601GUI0188",]
df2 <- df2[df2$LakeID != "1201ECA0697", ]
df2 <- df2[df2$LakeID != "1701LLO0708", ]
df2 <- df2[df2$LakeID != "1201GAL0563", ]

fit2<-lmer(log(TP_mg.L+.01) ~ Dreissenids + log(CA.SA) + Mean_Depth_m + (1|Sample_Year) + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=df2)

summary(fit2)

plot(fit2,col=df2$LakeID) #plots model residuals grouped by lake
plot(fit2,col=df2$LakeID, id=0.5, idLabels=~.obs, cex=.5) #plots model residuals with labels (observation number) for values outside Normal confidence limits 

plot(fit2,col=df2$Sample_Year) #plots model residuals grouped by year 
plot(fit2,col=df2$Sample_Year, id=0.5, idLabels=~.obs, cex=.5) #plots model residuals with labels (observation number) for values outside Normal confidence limits 

qqnorm(resid(fit2))
qqline(resid(fit2))
```

Removing lakes improves residual plot, but not QQ plot 

###Model 3 - removing "troublesome" Year 2015 
```{r}
#Select out data of interest
df3 <- df[df$Sample_Year != "2015",]


fit3<-lmer(log(TP_mg.L+.01) ~ Dreissenids + log(CA.SA) + Mean_Depth_m + (1|Sample_Year) + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=df3)

summary(fit3)

plot(fit3,col=df3$LakeID) #plots model residuals grouped by lake
plot(fit3,col=df3$LakeID, id=0.5, idLabels=~.obs, cex=.5) #plots model residuals with labels (observation number) for values outside Normal confidence limits 

plot(fit3,col=df3$Sample_Year) #plots model residuals grouped by year 
plot(fit3,col=df3$Sample_Year, id=0.5, idLabels=~.obs, cex=.5) #plots model residuals with labels (observation number) for values outside Normal confidence limits 

qqnorm(resid(fit3))
qqline(resid(fit3))
```

Removing 2015 improves neither residuals nor QQ 

###Model 3 - removing "troublesome" lakes AND 2015
```{r}
#Select out data of interest
df4 <- df[df$LakeID != "0601GUI0188",]
df4 <- df4[df4$LakeID != "1201ECA0697", ]
df4 <- df4[df4$LakeID != "1701LLO0708", ]
df4 <- df4[df4$LakeID != "1201GAL0563", ]
df4 <- df4[df4$Sample_Year != "2015",]
df4$LakeID <- levels(droplevels(df4$LakeID))

fit4<-lmer(log(TP_mg.L+.01) ~ Dreissenids + log(CA.SA) + Mean_Depth_m + (1|Sample_Year) + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=df4)

summary(fit4)

plot(fit4,col=df4$LakeID) #plots model residuals grouped by lake
plot(fit4,col=df4$LakeID, id=0.5, idLabels=~.obs, cex=.5) #plots model residuals with labels (observation number) for values outside Normal confidence limits 

plot(fit4,col=df4$Sample_Year) #plots model residuals grouped by year 
plot(fit4,col=df4$Sample_Year, id=0.5, idLabels=~.obs, cex=.5) #plots model residuals with labels (observation number) for values outside Normal confidence limits 

qqnorm(resid(fit4))
qqline(resid(fit4))

qqPlot(resid(fit4))

gg <- ggplot(data = df4, mapping = aes(sample = norm)) +
    geom_qq_band(bandType = "ks", mapping = aes(fill = "KS"), alpha = 0.5) +
    geom_qq_band(bandType = "ts", mapping = aes(fill = "TS"), alpha = 0.5) +
    geom_qq_band(bandType = "pointwise", mapping = aes(fill = "Normal"), alpha = 0.5) +
    geom_qq_band(bandType = "boot", mapping = aes(fill = "Bootstrap"), alpha = 0.5) +
    stat_qq_line() +
    stat_qq_point() +
    labs(x = "Theoretical Quantiles", y = "Sample Quantiles") +
    scale_fill_discrete("Bandtype")
gg
```

Removing "bad" lakes and Year 2015 improves residual plot the most, but the QQ line is still wonky 

##Alternate distributions using Generalized Linear Mixed Effects Model 

###Model 1 - Gaussian Family with Log Link

```{r}
#model transformed
Gfit<-glmer(TP_mg.L+1 ~ Dreissenids + log(CA.SA) + Mean_Depth_m + (1|Sample_Year) + (1 | Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=df4, family=gaussian(link="log"))

summary(Gfit)

plot(fitted(Gfit), residuals(Gfit), xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, lty = 2)
lines(smooth.spline(fitted(Gfit), residuals(Gfit)))

qqnorm(resid(Gfit))

simfit  <-  simulateResiduals(Gfit)
plotSimulatedResiduals(simfit) 
```

###Model 2 - Gamma with Inverse Link

```{r}
#model transformed
Gfit2<-glmer(TP_mg.L+.01 ~ Dreissenids + log(CA.SA) + Mean_Depth_m + (1|Sample_Year) + (1 | Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=df, family=Gamma(link="inverse"))

summary(Gfit2)

plot(fitted(Gfit2), residuals(Gfit2), xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, lty = 2)
lines(smooth.spline(fitted(Gfit2), residuals(Gfit2)))

```

###Model 2b - Gamma with Inverse Link and lakes and year 2015 removed

```{r}
#model transformed
Gfit2b<-glmer(TP_mg.L+.01 ~ Dreissenids + log(CA.SA) + Mean_Depth_m + (1|Sample_Year) + (1 | Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=df4, family=Gamma(link="inverse"))

summary(Gfit2b)

plot(fitted(Gfit2b), residuals(Gfit2b), xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, lty = 2)
lines(smooth.spline(fitted(Gfit2b), residuals(Gfit2b)))

```

###Model 3 - Gamma with Log Link
-Errors: 1) Model is nearly unidentifiable  2) Model failed to converge 
-Tried re-scaling Mean Depth using log transformation
```{r}
#model transformed
Gfit3<-glmer(TP_mg.L+.01 ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m) + (1|Sample_Year) + (1 | Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=df, family=Gamma(link="log"))

summary(Gfit3)

plot(fitted(Gfit3), residuals(Gfit3), xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, lty = 2)
lines(smooth.spline(fitted(Gfit3), residuals(Gfit3)))

simGfit3  <-  simulateResiduals(Gfit3)
plotSimulatedResiduals(simGfit3) 

```

###Model 3b - Gamma with Log Link and lakes and Year 2015 removed
-Error: 1) Model failed to converge 
```{r}
#model transformed
Gfit3b<-glmer(TP_mg.L+1 ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m) + (1|Sample_Year) + (1 | Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=df4, family=Gamma(link="log"))

summary(Gfit3b)

plot(fitted(Gfit3b), residuals(Gfit3b), xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, lty = 2)
lines(smooth.spline(fitted(Gfit3b), residuals(Gfit3b)))

simGfit3b  <-  simulateResiduals(Gfit3b)
plotSimulatedResiduals(simGfit3b) 

```

###Model 3c - Gamma with Identity Link

```{r}
#model transformed
Gfit3c<-glmer(TP_mg.L+1 ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m) + (1|Sample_Year) + (1 | Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=df, family=Gamma(link="identity"))

summary(Gfit3c)

plot(fitted(Gfit3c), residuals(Gfit3c), xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, lty = 2)
lines(smooth.spline(fitted(Gfit3c), residuals(Gfit3c)))

control<-glmerControl(optimizer="bobyqa")
simGfit3c  <-  simulateResiduals(Gfit3c)
plotSimulatedResiduals(simGfit3c) 

```

###Model 3d - Gamma with Identity Link and lakes/Sample Year 2015 removed
-Model failed to converge, model is nearly unidentifiable
```{r}
#model transformed
Gfit3d<-glmer(TP_mg.L+1.0 ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m)  + (1|LakeID) + (1|Sample_Year) + (1 | Sample_Month) + (1|Sample_Year:LakeID), data=df4, family=Gamma(link="identity"))

summary(Gfit3d)

plot(Gfit3d)

plot(fitted(Gfit3d), residuals(Gfit3d), xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, lty = 2)
lines(smooth.spline(fitted(Gfit3d), residuals(Gfit3d)))

qqPlot(resid(Gfit3d))

simGfit3d  <-  simulateResiduals(Gfit3d)
plotSimulatedResiduals(simGfit3d) 

```

###Model 3e - GLM (no random effects) Gamma with Identity Link and lakes/Sample Year 2015 removed

```{r}
#model transformed
Gfit3e<-glm(TP_mg.L+1 ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m), data=df4, family=Gamma(link="identity"))

summary(Gfit3e)

plot(Gfit3e)

plot(fitted(Gfit3e), residuals(Gfit3e), xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, lty = 2)
lines(smooth.spline(fitted(Gfit3e), residuals(Gfit3e)))


simGfit3e  <-  simulateResiduals(Gfit3e)
plotSimulatedResiduals(simGfit3e) 


```