---
title: "CSLAP MEMs (mixed) - June 2020"
author: "Victoria Field"
date: "6/8/2020"
output:
  html_document:
    toc: true
    toc_float: true
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(car)
library(lme4)
library(ggplot2)
require(lmerTest)
library(lattice)
library(predictmeans)
library(dplyr)
library(tidyr)
library(DHARMa)
library(magrittr)
library(pryr)
library(MASS)
```

#Data Read In 
```{r}
CSLAP<-read.csv("CSLAP_Dataset_09232019.csv", header=TRUE, stringsAsFactors = F)
CSLAP$Sample_Year<-as.factor(CSLAP$Sample_Year)
CSLAP$Sample_Month<-as.factor(CSLAP$Sample_Month)

#Create TN:TP column 
CSLAP$TN_TP<-(CSLAP$TN_mg.L+.01)/(CSLAP$TP_mg.L+.01)

#Read in and merge %ag 
Ag<-read.csv("Percent Ag Cover.csv", na.strings=c("", " "))
CSLAP<-merge(CSLAP, Ag, by="Lake_Name", all.x=TRUE)

#Split by `Info_Type`
OWCSLAP<-CSLAP[CSLAP$Info_Type == "OW",]
BSCSLAP<-CSLAP[CSLAP$Info_Type == "BS",]
SBCSLAP<-CSLAP[CSLAP$Info_Type == "SB",]
noSBCSLAP<-CSLAP[CSLAP$Info_Type != "SB",]

redCSLAP<-read.csv("redCSLAP.csv")
names(redCSLAP)[names(redCSLAP) == "Lake_Name.x"] <- "Lake_Name"
names(redCSLAP)[names(redCSLAP) == "Dreissenids.x"] <- "Dreissenids"
redCSLAP$Sample_Year<-as.factor(redCSLAP$Sample_Year)
redCSLAP$Sample_Month<-as.factor(redCSLAP$Sample_Month)

#Create TN:TP column 
redCSLAP$TN_TP<-(redCSLAP$TN_mg.L+.01)/(redCSLAP$TP_mg.L+.01)

#Merge %ag
redCSLAP<-merge(redCSLAP, Ag, by="Lake_Name", all.x=TRUE)

#Split by `Info_Type`
redOWCSLAP<-redCSLAP[redCSLAP$Info_Type == "OW",]
redBSCSLAP<-redCSLAP[redCSLAP$Info_Type == "BS",]
redSBCSLAP<-redCSLAP[redCSLAP$Info_Type == "SB",]
rednoSBCSLAP<-redCSLAP[redCSLAP$Info_Type != "SB",]
```

#Global Dataset 

##Open Water TN:TP - `lmer()`

```{r}
TNTP<-lmer(log(TN_TP+.01) ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m) + Percent_Ag +  (1|Sample_Year) + (1|Sample_Month) + (1|LakeID) + (1|Sample_Year:LakeID), data=OWCSLAP)

summary(TNTP)

plot(TNTP)

qqPlot(resid(TNTP))
```

##Open Water TN:TP - `glmer()`

```{r}
TNTP2<-glmer(TN_TP ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m) + Percent_Ag +  (1|LakeID) + (1|Sample_Year:LakeID), data=OWCSLAP, family=Gamma(link="log"))

summary(TNTP2)

plot(TNTP2)

qqPlot(resid(TNTP2))

simTNTP  <-  simulateResiduals(TNTP)
plotSimulatedResiduals(simTNTP) 
```

##Open Water Chlorophyll-a

-`Sample_Year` and `Sample_Month` removed for having low variance
```{r}
Chl<-lmer(log(Extracted_Chl.a_ug.L+1) ~ Dreissenids + log(CA.SA) + Mean_Depth_m + Percent_Ag  + (1|LakeID) + (1|Sample_Year:LakeID), data=OWCSLAP)

summary(Chl)

plot(Chl)

qqPlot(resid(Chl))
```

##Open Water Secchi Depth

-`Sample_Year` and `Sample_Month` removed for having low variance
```{r}
Secchi <- lmer(log(Secchi_Depth_m) ~ Dreissenids + log(CA.SA) + Mean_Depth_m + Percent_Ag + (1|LakeID) + (1|Sample_Year:LakeID), data=OWCSLAP)

summary(Secchi)

plot(Secchi)

qqPlot(resid(Secchi))
```

##Open Water True Color 

-Gamma model with no random effects comes out great, Gamma model with random effects much worse 

-Tried log-transforming True Color, but that made residuals and qqPlot worse

-`Sample_Year` and `Sample_Month` removed for having low variance
```{r}
OWTC<-lmer(True_Color_PTU ~ Dreissenids + log(CA.SA) + log(Mean_Depth_m) + Percent_Ag + (1|LakeID) + (1|Sample_Year:LakeID), data=OWCSLAP)

summary(OWTC)

plot(OWTC)

qqPlot(resid(OWTC))
```

##Shoreline Bloom Chlorophyll a 
-**update 03/08** `boundary (singular) fit` Removed random effect of Sample_Year with ~0.00 variance to resolve warning.

```{r}
SBChl<-lmer(log(ESF_Chl.a_ug.L) ~ Dreissenids + log(CA.SA) + Mean_Depth_m + Percent_Ag + (1|Sample_Month) + (1|LakeID) , data=SBCSLAP)

summary(SBChl)

plot(SBChl)

qqPlot(resid(SBChl))
```


##Shoreline Bloom Microcystin
 -**update 03/08** `boundary (singular) fit` Removed random effect of Sample_Month with 0.00 variance to resolve warning.
 
 -**update 06/05** adding continuous variable of TN:TP as a fixed effect
 
 -removed `(LakeID:Sample_Year)` because low variance

###Extract average TN:TP values for each lake and year
```{r, warning=FALSE, message=FALSE}
#Extracting average annual TP for each lake 
library(plyr)
avgTNTP<-ddply(OWCSLAP, c("LakeID", "Sample_Year"), summarize, 
             Mean = mean(TN_TP, na.rm=TRUE))
colnames(avgTNTP)[colnames(avgTNTP)=="Mean"] <- "TN_TP"

#Merge these values to the SBCSLAP df 
SBCSLAP<-SBCSLAP[,c(1:44, 46)]
SBCSLAP<-merge(SBCSLAP, avgTNTP, by=c("LakeID", "Sample_Year"), all.x=TRUE, all.y=FALSE)
```

```{r}
SBmicro<-lmer(log(ESF_Microcystin_ug.L) ~ Dreissenids + TN_TP + log(CA.SA) + Mean_Depth_m + Percent_Ag + (1|Sample_Year) + (1|LakeID), data=SBCSLAP)

summary(SBmicro)

plot(SBmicro)

qqPlot(resid(SBmicro))
```

#Reduced Dataset

##Open Water TN:TP 

-`Sample_Month` and `Sample_Year` removed for low variance 
```{r}
redTNTP<-glmer(TN_TP ~ Dreissenids + Percent_Ag +  (1|LakeID) + (1|Sample_Year:LakeID), data=redOWCSLAP, family=Gamma(link="log"))

summary(redTNTP)

plot(redTNTP)

qqPlot(resid(redTNTP))
```

##Open Water Chlorophyll-a

-`Sample_Year` and `Sample_Month` removed for having low variance
```{r}
redChl<-lmer(log(Extracted_Chl.a_ug.L+1) ~ Dreissenids + Percent_Ag  + (1|LakeID) + (1|Sample_Year:LakeID), data=redOWCSLAP)

summary(redChl)

plot(redChl)

qqPlot(resid(redChl))
```

##Open Water Secchi Depth

-`Sample_Year` and `Sample_Month` removed for having low variance
```{r}
redSecchi <- lmer(log(Secchi_Depth_m) ~ Dreissenids + Percent_Ag + (1|LakeID) + (1|Sample_Year:LakeID), data=redOWCSLAP)

summary(redSecchi)

plot(redSecchi)

qqPlot(resid(redSecchi))
```

##Open Water True Color 

-`Sample_Year` and `Sample_Month` removed for having low variance

-log-transforming True Color does not improve fit
```{r}
redOWTC<-lmer(True_Color_PTU ~ Dreissenids + Percent_Ag + (1|LakeID) + (1|Sample_Year:LakeID), data=redOWCSLAP)

summary(redOWTC)

plot(redOWTC)

qqPlot(resid(redOWTC))
```

##Shoreline Bloom Chlorophyll a 

-`Sample_Year` removed for low variance

```{r}
redSBChl<-lmer(log(ESF_Chl.a_ug.L) ~ Dreissenids  + Percent_Ag + (1|Sample_Month) + (1|LakeID) , data=redSBCSLAP)

summary(redSBChl)

plot(redSBChl)

qqPlot(resid(redSBChl))
```

##Shoreline Bloom Microcystin
 -**update 03/08** `boundary (singular) fit` Removed random effect of `Sample_Month` with 0.00 variance to resolve warning.
 
 -**update 06/05** adding continuous variable of TN:TP as a fixed effect
 
 -removed `(LakeID:Sample_Year)` because low variance

###Extract average TN:TP values for each lake and year
```{r, warning=FALSE}
#Extracting average annual TP for each lake 
library(plyr)
redavgTNTP<-ddply(redOWCSLAP, c("LakeID", "Sample_Year"), summarize, 
             Mean = mean(TN_TP, na.rm=TRUE))
colnames(redavgTNTP)[colnames(redavgTNTP)=="Mean"] <- "TN_TP"

#Merge these values to the SBCSLAP df 
redSBCSLAP<-redSBCSLAP[,c(1:52, 54)]
redSBCSLAP<-merge(redSBCSLAP, redavgTNTP, by=c("LakeID", "Sample_Year"), all.x=TRUE, all.y=FALSE)
```

```{r}
redSBmicro<-lmer(log(ESF_Microcystin_ug.L) ~ Dreissenids + TN_TP  + Percent_Ag + (1|Sample_Year) + (1|LakeID), data=redSBCSLAP)

summary(redSBmicro)

plot(redSBmicro)

qqPlot(resid(redSBmicro))
```

