---
title: "Historical CSLAP - Apr 29 2020"
author: "Victoria_Field"
date: "4/29/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lme4)
```

#Data Read In 
```{r}
#historicalCSLAP df edited to match DeRuyter and EatonBrook dfs
hisCSLAP<-read.csv("Historical_CSLAP_edited.csv", na.strings=c("", " ", "ND", "NA", "no filter", "nofilter"))

#DeRuyter and EatonBrook dfs
Eaton<-read.csv("EatonHistoricalClean2.csv", na.strings=c("", " ", "ND", "NA", "no filter", "nofilter"))
Der<-read.csv("DeruyterHistoricalClean2.csv", na.strings=c("", " ", "ND", "NA", "no filter", "nofilter"))

#change levels of Info_Type 
hisCSLAP$Info_Type[hisCSLAP$Info_Type == "ow"]<-"OW"
hisCSLAP$Info_Type[hisCSLAP$Info_Type == "sb"]<-"SB"
hisCSLAP$Info_Type<-droplevels(hisCSLAP$Info_Type)
```

#Merge hisCSLAP with Deruyter and Eatonbrook 
```{r}
#remove any observations from Eaton or DeRuyter from hisCSLAP 
hisCSLAP<-hisCSLAP[hisCSLAP$Lake_Name != "De Ruyter Reservoir",]
hisCSLAP<-hisCSLAP[hisCSLAP$Lake_Name != "Eaton Brook Reservoir",]

#row bind cleaned DeRuyter and EatonBrook 
hisCSLAP<-rbind(hisCSLAP, Eaton)
hisCSLAP<-rbind(hisCSLAP, Der)
```

#Further organize hisCSLAP, and merge physical characteristics and invasion status 
```{r}
#extract sample year and month 
hisCSLAP$Sample_Year<-as.numeric(format(as.Date(hisCSLAP$Sample_Date, format="%m/%d/%Y"),"%Y"))
hisCSLAP$Sample_Month<-as.numeric(format(as.Date(hisCSLAP$Sample_Date, format="%m/%d/%Y"),"%m"))

#read in CA:SA, Mean_Depths, and Year of Invasion
CA_SA<-read.csv("Catchment_SA.csv", header=TRUE, colClasses=c("character", "numeric"))
CA_SA[13,1]<- "De Ruyter Reservoir"

Mean_Depth<-read.csv("MeanDepths.csv", header=TRUE, colClasses=c("character", "numeric"))

Invasion_Year<-read.csv("InvasionYear.csv", header=TRUE, colClasses=c("factor", "numeric"))

#merge with hisCSLAP
hisCSLAP<-merge(hisCSLAP, CA_SA, all.x=TRUE)
hisCSLAP<-merge(hisCSLAP, Mean_Depth, all.x=TRUE)
hisCSLAP<-merge(hisCSLAP, Invasion_Year, all.x=TRUE)

#create new column `Years_Invaded` by subtracting `Invasion Year` from `Sample Year`
hisCSLAP$Years_Invaded<-hisCSLAP$Sample_Year-hisCSLAP$Invasion_Year

#turn `Sample_Year`, `Sample_Month`, and `Invasion_Year` back to factors
hisCSLAP$Sample_Year<-as.factor(hisCSLAP$Sample_Year)
hisCSLAP$Sample_Month<-as.factor(hisCSLAP$Sample_Month)
hisCSLAP$Invasion_Year<-as.factor(hisCSLAP$Invasion_Year)

#create column with 'Invaded' or "Uninvaded" based on `Years Invaded`
hisCSLAP$Dreissenids<-ifelse(hisCSLAP$Years_Invaded >= 0, "Invaded", "Uninvaded")
hisCSLAP$Dreissenids<-as.factor(hisCSLAP$Dreissenids)

#we will also remove and shoreline bloom samples 
hisCSLAP<-hisCSLAP[hisCSLAP$Info_Type != "SB",]
hisCSLAP$Info_Type<-droplevels(hisCSLAP$Info_Type)

#finally, create a df with only OW samples and a df with only BS samples 
OWhisCSLAP<-hisCSLAP[hisCSLAP$Info_Type == "OW",]
BShisCSLAP<-hisCSLAP[hisCSLAP$Info_Type == "BS",]
```

#Aggregating yearly means by lake and years invaded
```{r}
Chl<-aggregate(Extracted_Chl_ug.L ~ Lake_Name + Years_Invaded, data=hisCSLAP[hisCSLAP$Info_Type == "OW",], FUN=mean)

OWTP<-aggregate(TP_mg.L ~ Lake_Name + Lake_Name + Years_Invaded, data=hisCSLAP[hisCSLAP$Info_Type == "OW",], FUN=mean)

BSTP<-aggregate(TP_mg.L ~ Lake_Name + Lake_Name + Years_Invaded, data=hisCSLAP[hisCSLAP$Info_Type == "BS",], FUN=mean)

OWTN<-aggregate(TN_mg.L ~ Lake_Name + Lake_Name + Years_Invaded, data=hisCSLAP[hisCSLAP$Info_Type == "OW",], FUN=mean)

TC<-aggregate(True_Color_PTU ~ Lake_Name + Years_Invaded, data=hisCSLAP[hisCSLAP$Info_Type == "OW",], FUN=mean)

Calc<-aggregate(Calcium_mg.L ~ Lake_Name + Years_Invaded, data=hisCSLAP[hisCSLAP$Info_Type == "OW",], FUN=mean)
```

#How do lakes change after invasion? (t-test for pre and post invasion, then MEMs)

##Extracted Chlorophyll-a
```{r}
hist(log(OWhisCSLAP$Extracted_Chl_ug.L))

t.test(Extracted_Chl_ug.L ~ Dreissenids, data=hisCSLAP[hisCSLAP$Info_Type == "OW",])

Chl.lmer<-lmer(log(Extracted_Chl_ug.L) ~ Dreissenids + (1|LakeID), data=hisCSLAP[hisCSLAP$Info_Type == "OW",])

summary(Chl.lmer)

plot(Chl.lmer)
```

##Open Water TP
```{r}
hist(log(OWhisCSLAP$TP_mg.L))

t.test(TP_mg.L ~ Dreissenids, data=hisCSLAP[hisCSLAP$Info_Type == "OW",])

OWTP.lmer<-lmer(log(TP_mg.L) ~ Dreissenids  + (1|LakeID), data=hisCSLAP[hisCSLAP$Info_Type == "OW",])

summary(OWTP.lmer)

plot(OWTP.lmer)
```

##Bottom Sample TP
```{r}
hist(log(BShisCSLAP$TP_mg.L))

t.test(TP_mg.L ~ Dreissenids, data=hisCSLAP[hisCSLAP$Info_Type == "BS",])

BSTP.lmer<-lmer(log(TP_mg.L) ~ Dreissenids + (1|LakeID), data=hisCSLAP[hisCSLAP$Info_Type == "BS",])

summary(BSTP.lmer)

plot(BSTP.lmer)
```

##Open Water TN
```{r}
hist(log(OWhisCSLAP$TN_mg.L))

t.test(TN_mg.L ~ Dreissenids, data=hisCSLAP[hisCSLAP$Info_Type == "OW",])

OWTN.lmer<-lmer(log(TN_mg.L) ~ Dreissenids + (1|LakeID), data=hisCSLAP[hisCSLAP$Info_Type == "OW",])

summary(OWTN.lmer)
```

##Open Water True Color
```{r}
t.test(True_Color_PTU ~ Dreissenids, data=hisCSLAP[hisCSLAP$Info_Type == "OW",])

TC.lmer<-lmer(True_Color_PTU ~ Dreissenids + (1|LakeID), data=hisCSLAP[hisCSLAP$Info_Type == "OW",])

summary(TC.lmer)
```

##Calcium
```{r}
t.test(Calcium_mg.L ~ Dreissenids, data=hisCSLAP[hisCSLAP$Info_Type == "OW",])

Calc.lmer<-lmer(Calcium_mg.L ~ Dreissenids + (1|LakeID), data=hisCSLAP[hisCSLAP$Info_Type == "OW",])

summary(Calc.lmer)
```

#Boxplots
##Chlorophyll-a
```{r}
ggplot(hisCSLAP[hisCSLAP$Info_Type == "OW",], aes(x=Dreissenids, y=Extracted_Chl_ug.L)) + 
  geom_boxplot()+
  theme_bw()+
  labs(y = "Extracted Chlorophyll-a (mg/L)")

ggsave("historical_Chl.png")
```

##OWTP
```{r}
ggplot(hisCSLAP[hisCSLAP$Info_Type == "OW",], aes(x=Dreissenids, y=TP_mg.L)) + 
  geom_boxplot()+
  theme_bw()+
  labs(y = "Open Water TP (mg/L)")

ggsave("historical_OWTP.png")
```

##BSTP
```{r}
ggplot(hisCSLAP[hisCSLAP$Info_Type == "BS",], aes(x=Dreissenids, y=TP_mg.L)) + 
  geom_boxplot()+
  theme_bw()+
  labs(y = "Bottom Sample TP (mg/L)")

ggsave("historical_BSTP.png")
```

##OWTN
```{r}
ggplot(hisCSLAP[hisCSLAP$Info_Type == "OW",], aes(x=Dreissenids, y=TN_mg.L)) + 
  geom_boxplot()+
  theme_bw()+
  labs(y = "Open Water TN (mg/L)")

ggsave("historical_OWTN.png")
```